---
title: "Report Peak Finding Merck V2"
subtitle: "C++ Implementation and Integration in StreamFind"
author: "Ricardo Cunha"
date: "`r format(Sys.time(), '%d %B, %Y')`"
toc: true
format:
  html:
    grid:
      sidebar-width: 300px
      body-width: 1200px
      margin-width: 300px
      gutter-width: 1.5rem
    other-links:
      - text: StreamFind
        href: https://odea-project.github.io/StreamFind/
        icon: github
    html-math-method: katex
    theme: journal
    code-tools: true
    anchor-sections: true
    code-fold: true
    code-overflow: scroll
    code-line-numbers: false
    code-copy: true
    cap-location: top
    self-contained: true
    embed-resources: true
    toc: true
    toc-location: left
execute:
  warning: false
---

```{r}
#| include: false

options(DT.options = list(
  dom = "lfrtip",
  autoWidth = TRUE,
  pageLength = 10,
  lengthMenu = c(10, 20, 50, 75, 100),
  initComplete = htmlwidgets::JS(
    "function(settings, json) {",
    "$(this.api().table().body()).css({'font-size': '12px'});",
    "$(this.api().table().body()).css({'white-space': 'nowrap'});",
    "}"
  )
))

simple_table_opt <- list(
  dom = 't',
  paging = FALSE,
  searching = FALSE,
  info = FALSE
)

library(knitr)
library(kableExtra)
library(data.table)
library(DT)
library(magrittr)
library(ggplot2)
library(plotly)
library(StreamFind)

devtools::load_all()

source(file.path("C:\\Users\\apoli\\Documents\\github\\StreamFind\\dev\\merck_peak_finding\\dev_resources.R"))

examples <- unique(files_merck_ex$example)
```

# Files

```{r}
#| label: all-files

DT::datatable(
  files_merck_ex[, 1:3],
  options = simple_table_opt
)
```

# Targets

```{r}
#| label: targets-merck

DT::datatable(db_merck_ex)
```

# Peak Finding Targets

```{r}
#| label: peak-finding-targets-merck-function
#| include: false

run_peak_finding_debug_targets <- function(example_name) {
  db_merck_ex1 <- db_merck_ex[db_merck_ex$example %in% example_name, ]
  files <- files_merck_ex$file_path[files_merck_ex$example %in% example_name & !files_merck_ex$blank]
  
  for (i in seq_len(nrow(db_merck_ex1))) {
    cat(
      "Target ", i, ": ", db_merck_ex1$name[i],
      " - m/z: ", db_merck_ex1$mz[i],
      " - rt: ", db_merck_ex1$rt[i], "\n"
    )
    ps_ff <- MassSpecMethod_FindFeatures_native(
      rtWindows = data.frame(rtmin = 300, rtmax = 3400),
      resolution_profile = c(25000L, 55000L, 80000L),
      noiseThreshold = 15,
      minSNR = 3,
      minTraces = 3,
      baselineWindow = 200,
      maxWidth = 250, #100
      base_quantile = 0.99,
      debug_mz = db_merck_ex1$mz[i]
    )
    clear_cache("all")
    ms <- MassSpecEngine$new(analyses = files)
    ms$Workflow = list(ps_ff)
    ms$run_workflow()
  }
  log_path <- "C:\\Users\\apoli\\Documents\\github\\StreamFind\\log"
  log_files <- list.files(log_path, pattern = "peak_detection", full.names = TRUE)
  folder_ex1 <- file.path(log_path, paste0("example_", example_name))
  if (!dir.exists(folder_ex1)) {
    dir.create(folder_ex1)
  }
  for (log_file in log_files) {
    file.copy(log_file, folder_ex1, overwrite = TRUE)
  }
  if (all(file.exists(log_files))) {
    cat("All log files copied successfully to", folder_ex1, "\n")
  } else {
    cat("Some log files were not copied successfully.\n")
  }
  for (log_file in log_files) {
    file.remove(log_file)
  }
}
```

```{r}
#| label: run-peak-finding-debug-merck
#| include: false

#temp <- lapply(examples, run_peak_finding_debug_targets)
#run_peak_finding_debug_targets(examples[2])
```

```{r}
#| label: run-peak-finding-debug-merck-single
#| include: false

# example_name <- examples[2]
# db_merck_ex1 <- db_merck_ex[db_merck_ex$example %in% example_name, ]
# files <- files_merck_ex$file_path[files_merck_ex$example %in% example_name & !files_merck_ex$blank]
# ps_ff <- MassSpecMethod_FindFeatures_native(
#   rtWindows = data.frame(rtmin = 300, rtmax = 3400),
#   resolution_profile = c(25000L, 55000L, 80000L),
#   noiseThreshold = 15,
#   minSNR = 3,
#   minTraces = 3,
#   baselineWindow = 200,
#   maxWidth = 250, #100
#   base_quantile = 0.99,
#   debug_mz = db_merck_ex1$mz[4]
# )
# clear_cache("all")
# ms <- MassSpecEngine$new(analyses = files)
# ms$Workflow = list(ps_ff)
# ms$run_workflow()
```

```{r}
#| label: debug-plots
#| results: hide
#| include: false

source("C:\\Users\\apoli\\Documents\\github\\StreamFind\\dev\\merck_peak_finding\\dev_log_plot.R")
log_path <- "C:\\Users\\apoli\\Documents\\github\\StreamFind\\log"
log_example_folders <- list.dirs(log_path, recursive = FALSE, full.names = TRUE)
log_example_folders <- log_example_folders[grepl("example_", log_example_folders)]

plot_list <- list()

for (folder in log_example_folders) {
  log_files <- list.files(folder, pattern = "peak_detection", full.names = TRUE)
  example <- gsub(basename(folder), pattern = "example_", replacement = "")
  plot_list[[example]] <- list()
  for (log_file in log_files) {
    tryCatch({
      plot_temp <- plot_cluster_data(plot3D = FALSE, log_file = log_file)
      log_file_name <- basename(log_file)
      plot_list[[example]][[log_file_name]] <- plot_temp
    }, error = function(e) {
      cat("Error processing", log_file, ":", conditionMessage(e), "\n")
    })
  }
}
```

## Example `r examples[1]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-1
#| output: asis
#| echo: false

example <- examples[1]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[2]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-2
#| output: asis
#| echo: false

example <- examples[2]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[3]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-3
#| output: asis
#| echo: false

example <- examples[3]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[4]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-4
#| output: asis
#| echo: false

example <- examples[4]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

# Comparison with Other Algorithms

The C++ native peak finding is compared against OpenMS and XCMS3 (centWave).

```{r}
#| label: create-engine

engine <- MassSpecEngine$new(
  analyses = files_merck_ex$file_path[!files_merck_ex$blank]
)

engine
```

## Algorithms

```{r}
#| label: ps-ff-definitions

ps_ff_sf <- MassSpecMethod_FindFeatures_native(
  rtWindows = data.frame(rtmin = 300, rtmax = 3400),
  resolution_profile = c(25000L, 55000L, 80000L),
  noiseThreshold = 15,
  minSNR = 3,
  minTraces = 3,
  baselineWindow = 200,
  maxWidth = 250,
  base_quantile = 0.99,
  debug_mz = 0
)

ps_ff_openms <- MassSpecMethod_FindFeatures_openms(
  noiseThrInt = 15,         # Aligned with native noiseThreshold
  chromSNR = 3,             # Aligned with native minSNR
  chromFWHM = 7,
  mzPPM = 15,
  reEstimateMTSD = TRUE,
  traceTermCriterion = "sample_rate",
  traceTermOutliers = 5,
  minSampleRate = 1,
  minTraceLength = 3,       # Aligned with native minTraces (approximate - native uses count, openms uses seconds)
  maxTraceLength = 250,     # Aligned with native maxWidth
  widthFiltering = "fixed",
  minFWHM = 4,
  maxFWHM = 250,            # Aligned with native maxWidth
  traceSNRFiltering = TRUE,
  localRTRange = 0,
  localMZRange = 0,
  isotopeFilteringModel = "none",
  MZScoring13C = FALSE,
  useSmoothedInts = FALSE,
  intSearchRTWindow = 3,
  useFFMIntensities = FALSE,
  verbose = FALSE
)

ps_ff_xcms <- MassSpecMethod_FindFeatures_xcms3_centwave(
  ppm = 15,                 # Aligned with native (uses resolution_profile, ~15 ppm)
  peakwidth = c(5, 250),    # Aligned with native maxWidth (250 seconds)
  snthresh = 3,             # Aligned with native minSNR
  prefilter = c(3, 15),     # Aligned with native minTraces (3) and noiseThreshold (15)
  mzCenterFun = "wMean",
  integrate = 1,
  mzdiff = -2e-04,
  fitgauss = TRUE,
  noise = 15,               # Aligned with native noiseThreshold
  verboseColumns = TRUE,
  firstBaselineCheck = FALSE,
  extendLengthMSW = FALSE
)
```

```{r}
#| label: run-speed-test
#| include: false

#source("C:\\Users\\apoli\\Documents\\github\\StreamFind\\dev\\merck_peak_finding\\run_speed_test.R")
results_dir <- "C:\\Users\\apoli\\Documents\\github\\StreamFind\\dev\\merck_peak_finding"
all_results <- fread(file.path(results_dir, "speed_test_all_results.csv"))
summary_stats <- fread(file.path(results_dir, "speed_test_summary_stats.csv"))
all_intensity_data <- fread(file.path(results_dir, "speed_test_intensity_data.csv"))
intensity_summary <- fread(file.path(results_dir, "speed_test_intensity_summary.csv"))
```

## Speed Test Summary

```{r}
#| label: display-speed-test-summary

# Display summary statistics
DT::datatable(
  summary_stats,
  options = list(
    dom = 't',
    paging = FALSE
  ),
  caption = "Speed Test - Summary Statistics"
) %>%
  DT::formatRound(columns = c("mean_time", "sd_time", "features_per_second"), digits = 2) %>%
  DT::formatRound(columns = c("mean_features"), digits = 0)
```

## Intensity Distribution Comparison

```{r}
#| label: intensity-distribution-histograms
#| results: hide

# Create interactive plotly histogram data for comparison
if (!is.null(all_intensity_data) && nrow(all_intensity_data) > 0) {
  # Calculate histogram bins for each algorithm
  n_bins <- 50
  
  # Create color palette
  colors <- c("#F8766D", "#00BA38", "#619CFF")
  algorithms <- unique(all_intensity_data$algorithm)
  
  # Histogram on log10 scale (Plotly)
  p1 <- plot_ly() %>%
    layout(
      title = "Feature Intensity Distribution (Log10 Scale)",
      xaxis = list(title = "Log10(Intensity + 1)"),
      yaxis = list(title = "Count"),
      barmode = "overlay",
      legend = list(x = 0.5, y = -0.2, orientation = "h", xanchor = "center")
    )
  
  for (i in seq_along(algorithms)) {
    alg <- algorithms[i]
    data_subset <- all_intensity_data[algorithm == alg, log10_intensity]
    p1 <- p1 %>%
      add_histogram(
        x = data_subset,
        name = alg,
        opacity = 0.6,
        marker = list(color = colors[i]),
        nbinsx = n_bins
      )
  }
  
  # Density plot on log10 scale (Plotly)
  p2 <- plot_ly() %>%
    layout(
      title = "Feature Intensity Density (Log10 Scale)",
      xaxis = list(title = "Log10(Intensity + 1)"),
      yaxis = list(title = "Density"),
      legend = list(x = 0.5, y = -0.2, orientation = "h", xanchor = "center")
    )
  
  for (i in seq_along(algorithms)) {
    alg <- algorithms[i]
    data_subset <- all_intensity_data[algorithm == alg, log10_intensity]
    density_result <- density(data_subset, n = 512)
    p2 <- p2 %>%
      add_lines(
        x = density_result$x,
        y = density_result$y,
        name = alg,
        fill = "tozeroy",
        fillcolor = paste0(colors[i], "4D"),
        line = list(color = colors[i], width = 2)
      )
  }
  
  # Box plot comparison - Log10 scale (Plotly)
  p3_log <- plot_ly() %>%
    layout(
      title = "Feature Intensity Distribution by Algorithm (Log10 Scale)",
      xaxis = list(title = "Algorithm"),
      yaxis = list(title = "Log10(Intensity + 1)"),
      legend = list(x = 0.5, y = -0.2, orientation = "h", xanchor = "center")
    )
  
  for (i in seq_along(algorithms)) {
    alg <- algorithms[i]
    data_subset <- all_intensity_data[algorithm == alg, log10_intensity]
    p3_log <- p3_log %>%
      add_boxplot(
        y = data_subset,
        name = alg,
        fillcolor = paste0(colors[i], "4D"),
        marker = list(
          color = colors[i],
          outliercolor = colors[i]
        ),
        line = list(color = colors[i]),
        boxmean = TRUE
      )
  }
  
  # Box plot comparison - Actual intensity scale (Plotly)
  p3_actual <- plot_ly() %>%
    layout(
      title = "Feature Intensity Distribution by Algorithm",
      xaxis = list(title = "Algorithm"),
      yaxis = list(title = "Intensity"),
      legend = list(x = 0.5, y = -0.2, orientation = "h", xanchor = "center")
    )
  
  for (i in seq_along(algorithms)) {
    alg <- algorithms[i]
    data_subset <- all_intensity_data[algorithm == alg, intensity]
    p3_actual <- p3_actual %>%
      add_boxplot(
        y = data_subset,
        name = alg,
        fillcolor = paste0(colors[i], "4D"),
        marker = list(
          color = colors[i],
          outliercolor = colors[i]
        ),
        line = list(color = colors[i]),
        boxmean = TRUE
      )
  }
  
} else {
  cat("No intensity data available for comparison.\n")
}
```

```{r}
#| label: plot-intensity-distribution-histograms
#| fig-width: 12
#| fig-height: 8

p1
```

```{r}
#| label: plot-intensity-distribution-density
#| fig-width: 12
#| fig-height: 8

p2
```

```{r}
#| label: plot-intensity-distribution-boxplot-log
#| fig-width: 12
#| fig-height: 8

p3_log
```

```{r}
#| label: plot-intensity-distribution-boxplot-actual
#| fig-width: 12
#| fig-height: 8

p3_actual
```

```{r}
#| label: display-intensity-summary

# Display intensity distribution summary table
if (!is.null(intensity_summary) && nrow(intensity_summary) > 0) {
  DT::datatable(
    intensity_summary,
    options = list(
      dom = 't',
      paging = FALSE
    ),
    caption = "Intensity Distribution Summary by Algorithm"
  ) %>%
    DT::formatRound(columns = c("mean_intensity", "median_intensity", "sd_intensity", 
                                  "q25_intensity", "q75_intensity", "min_intensity", 
                                  "max_intensity"), digits = 0)
} else {
  cat("No intensity summary data available.\n")
}
```

# StreamFind Workflow

```{r}
demo <- MassSpecEngine$new(
  analyses = files_merck_ex$file_path[c(1, 2)],
  metadata = list(name = "Demo Project")
)

demo$Analyses <- set_replicate_names(demo$Analyses, c("blank", "sample"))
demo$Analyses <- set_blank_names(demo$Analyses, rep("blank", 2))
demo
```

## Inspect TIC

```{r}
#| label: plot-demo-tic

plot_spectra_tic(demo$Analyses, downsize = 5, levels = 1)
```

## Workflow

```{r}
#| label: run-demo-workflow

workflow <- list(
  MassSpecMethod_FindFeatures_native(
    rtWindows = data.frame(rtmin = 300, rtmax = 3400),
    resolution_profile = c(25000L, 55000L, 80000L),
    noiseThreshold = 15,
    minSNR = 3,
    minTraces = 3,
    baselineWindow = 200,
    maxWidth = 250,
    base_quantile = 0.99,
    debug_mz = 0
  )
)

demo$Workflow <- workflow
demo
```

```{r}
#| label: run-demo-workflow-execute

demo$run_workflow()
```

## Results

```{r}
#| label: show-demo-results
#| fig-height: 8

plot_features(
  demo$Results[[1]],
  analyses = 2,
  mz = db_merck_ex[db_merck_ex$example %in% examples[1], ],
  ppm = 20,
  sec = 60
)
```
