---
title: "SF2025_01 - Report Peak Finding Merck (Milestone 1, part 2/2)"
subtitle: "C++ Implementation and Integration in StreamFind"
author: "Ricardo Cunha"
date: "`r format(Sys.time(), '%d %B, %Y')`"
toc: true
format:
  html:
    grid:
      sidebar-width: 300px
      body-width: 1200px
      margin-width: 300px
      gutter-width: 1.5rem
    other-links:
      - text: StreamFind
        href: https://odea-project.github.io/StreamFind/
        icon: github
    html-math-method: katex
    theme: journal
    code-tools: true
    anchor-sections: true
    code-fold: true
    code-overflow: scroll
    code-line-numbers: false
    code-copy: true
    cap-location: top
    self-contained: true
    embed-resources: true
    toc: true
    toc-location: left
execute:
  warning: false
---

```{r}
#| include: false

options(DT.options = list(
  dom = "lfrtip",
  autoWidth = TRUE,
  pageLength = 10,
  lengthMenu = c(10, 20, 50, 75, 100),
  initComplete = htmlwidgets::JS(
    "function(settings, json) {",
    "$(this.api().table().body()).css({'font-size': '12px'});",
    "$(this.api().table().body()).css({'white-space': 'nowrap'});",
    "}"
  )
))

simple_table_opt <- list(
  dom = 't',
  paging = FALSE,
  searching = FALSE,
  info = FALSE
)

library(knitr)
library(kableExtra)
library(data.table)
library(DT)
library(magrittr)
library(ggplot2)
library(plotly)
library(StreamFind)

devtools::load_all()

source(file.path("C:\\Users\\apoli\\Documents\\github\\StreamFind\\dev\\merck_peak_finding\\dev_resources.R"))

examples <- unique(merck2_files_ex$example)
```

# Demonstration - Milestone 1

A total of 10 datasets were provided by Merck for testing and validation of the peak finding algorithm. The focus is on the sample measurements not yet the associated blank runs. For each dataset, a set of targets with expected m/z and retention time values were provided for validation of the peak finding results.

```{r}
#| label: all-files

DT::datatable(merck2_files_ex[, 1:3])
```

# Targets

```{r}
#| label: targets-merck

DT::datatable(merck2_db_ex)
```

# Peak Finding Targets

```{r}
#| label: peak-finding-targets-merck-function
#| include: false

run_peak_finding_debug_targets <- function(example_name) {
  merck2_db_ex1 <- merck2_db_ex[merck2_db_ex$example %in% example_name, ]
  files <- merck2_files_ex$file_path[merck2_files_ex$example %in% example_name & !merck2_files_ex$blank]
  
  for (i in seq_len(nrow(merck2_db_ex1))) {
    cat(
      "Target ", i, ": ", merck2_db_ex1$name[i],
      " - m/z: ", merck2_db_ex1$mz[i],
      " - rt: ", merck2_db_ex1$rt[i], "\n"
    )
    ps_ff <- DB_MassSpecMethod_FindFeatures_native(
      rtWindows = data.frame(rtmin = 300, rtmax = 3400),
      resolution_profile = c(25000L, 55000L, 80000L),
      noiseThreshold = 15,
      minSNR = 3,
      minTraces = 3,
      baselineWindow = 400,
      maxWidth = 250,
      base_quantile = 0.99,
      debug_mz = merck2_db_ex1$mz[i]
    )
    clear_cache("all")
    prj <- file.path("dev", "merck_peak_finding", "data")
    ms <- DB_MassSpecEngine$new(
      project_path = prj,
      files = files
    )
    ms$clear_cache()
    ms$Workflow = list(ps_ff)
    ms$run_workflow()
  }
  log_path <- "C:\\Users\\apoli\\Documents\\github\\StreamFind\\log"
  log_files <- list.files(log_path, pattern = "peak_detection", full.names = TRUE)
  folder_ex1 <- file.path(log_path, paste0("example_", example_name))
  if (!dir.exists(folder_ex1)) {
    dir.create(folder_ex1)
  }
  for (log_file in log_files) {
    file.copy(log_file, folder_ex1, overwrite = TRUE)
  }
  if (all(file.exists(log_files))) {
    cat("All log files copied successfully to", folder_ex1, "\n")
  } else {
    cat("Some log files were not copied successfully.\n")
  }
  for (log_file in log_files) {
    file.remove(log_file)
  }
  file.remove(list.files(prj, full.names = TRUE))
}
```

```{r}
#| label: run-peak-finding-debug-merck
#| include: false

#temp <- lapply(examples, run_peak_finding_debug_targets)
```

```{r}
#| label: debug-plots
#| results: hide
#| include: false

source("C:\\Users\\apoli\\Documents\\github\\StreamFind\\dev\\merck_peak_finding\\dev_log_plot.R")
log_path <- "C:\\Users\\apoli\\Documents\\github\\StreamFind\\log"
log_example_folders <- list.dirs(log_path, recursive = FALSE, full.names = TRUE)
log_example_folders <- log_example_folders[grepl("example_", log_example_folders)]

plot_list <- list()

for (folder in log_example_folders) {
  log_files <- list.files(folder, pattern = "peak_detection", full.names = TRUE)
  example <- gsub(basename(folder), pattern = "example_", replacement = "")
  plot_list[[example]] <- list()
  for (log_file in log_files) {
    tryCatch({
      plot_temp <- plot_cluster_data(plot3D = FALSE, log_file = log_file)
      log_file_name <- basename(log_file)
      plot_list[[example]][[log_file_name]] <- plot_temp
    }, error = function(e) {
      cat("Error processing", log_file, ":", conditionMessage(e), "\n")
    })
  }
}
```

## Example `r examples[1]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-1
#| output: asis
#| echo: false

example <- examples[1]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[2]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-2
#| output: asis
#| echo: false

example <- examples[2]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[3]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-3
#| output: asis
#| echo: false

example <- examples[3]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[4]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-4
#| output: asis
#| echo: false

example <- examples[4]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[5]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-5
#| output: asis
#| echo: false

example <- examples[5]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[6]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-6
#| output: asis
#| echo: false

example <- examples[6]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[7]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-7
#| output: asis
#| echo: false

example <- examples[7]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[8]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-8
#| output: asis
#| echo: false

example <- examples[8]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[9]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-9
#| output: asis
#| echo: false

example <- examples[9]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Example `r examples[10]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-10
#| output: asis
#| echo: false

example <- examples[10]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

```{r}
#| include: false

# ps_ff <- DB_MassSpecMethod_FindFeatures_native(
#   rtWindows = data.frame(rtmin = 300, rtmax = 3400),
#   resolution_profile = c(25000L, 55000L, 80000L),
#   noiseThreshold = 15,
#   minSNR = 3,
#   minTraces = 3,
#   baselineWindow = 400,
#   maxWidth = 250,
#   base_quantile = 0.99,
#   debug_mz = 0
# )

# clear_cache("all")
# prj <- file.path("dev", "merck_peak_finding", "data")

# ms <- DB_MassSpecEngine$new(
#   project_path = prj,
#   files = merck2_files_ex$file_path[merck2_files_ex$example %in% examples[8] & !merck2_files_ex$blank]
# )

# ms$clear_cache()
# ms$Workflow = list(ps_ff)
# ms$run_workflow()

# plot_features(
#   ms$NonTargetAnalysis,
#   mz = data.frame(mzmin = 790, mzmax = 820, rtmin = 20 * 60, rtmax = 22 * 60)
# )

# file.remove(list.files(prj, full.names = TRUE))
```
