---
title: "SF2025_01 - Report Peak Finding Merck (Milestone 1, part 1/2)"
subtitle: "C++ Implementation and Integration in StreamFind"
author: "Ricardo Cunha"
date: "`r format(Sys.time(), '%d %B, %Y')`"
toc: true
format:
  html:
    grid:
      sidebar-width: 300px
      body-width: 1200px
      margin-width: 300px
      gutter-width: 1.5rem
    other-links:
      - text: StreamFind
        href: https://odea-project.github.io/StreamFind/
        icon: github
    html-math-method: katex
    theme: journal
    code-tools: true
    anchor-sections: true
    code-fold: true
    code-overflow: scroll
    code-line-numbers: false
    code-copy: true
    cap-location: top
    self-contained: true
    embed-resources: true
    toc: true
    toc-location: left
execute:
  warning: false
---

```{r}
#| include: false

options(DT.options = list(
  dom = "lfrtip",
  autoWidth = TRUE,
  pageLength = 10,
  lengthMenu = c(10, 20, 50, 75, 100),
  initComplete = htmlwidgets::JS(
    "function(settings, json) {",
    "$(this.api().table().body()).css({'font-size': '12px'});",
    "$(this.api().table().body()).css({'white-space': 'nowrap'});",
    "}"
  )
))

simple_table_opt <- list(
  dom = 't',
  paging = FALSE,
  searching = FALSE,
  info = FALSE
)

library(knitr)
library(kableExtra)
library(data.table)
library(DT)
library(magrittr)
library(ggplot2)
library(plotly)
library(StreamFind)

devtools::load_all()

source(file.path("C:\\Users\\apoli\\Documents\\github\\StreamFind\\dev\\merck_peak_finding\\dev_resources.R"))

examples <- unique(files_merck_ex$example)
```

# Merck Peak Finding - Milestone 1

This report described the results of the first phase of the low intenisty peak algorithm for Merck. This phase focused on the development of an algorithm for detecting low-intensity peaks using deterministic approaches and machine learning methods. No clear advantage of machine learning methods was observed in this phase, and therefore the focus shifted to optimizing the deterministic C++ implementation integrated into StreamFind. The report includes the overview of the algorithm logic, demonstration of the C++ implementation using four curated datasets from Merck, and a comparative analysis of its performance against other established algorithms such as OpenMS and XCMS3 (centWave). The evaluation is based on precision and recall metrics using manually curated peak lists from Merck as a reference. The second part of the report includes additional tests based on another curated datasets from Merck.

## Algorithm Logic

### Overview

The algorithm was first developed in R and then implemented in C++ for better performance and integration into StreamFind. The  `nts::NTS_DATA::find_features()` function implements a comprehensive, deterministic peak detection algorithm optimized for low-intensity (i.e., 200 to 1000 counts) mass spectrometry data. Yet, high-intensity peaks (i.e., above 8M counts) are also detected effectively. The algorithm processes liquid chromatography-mass spectrometry (LC-MS) data through a sequential pipeline of noise reduction, m/z clustering, baseline estimation, and derivative-based peak detection, with separate processing for positive and negative ion modes. The raw code can be found in the `src/nts/nts.cpp` and `src/nts/utils.cpp` files of the branch file_system of the StreamFind repository.

### Algorithm Steps

#### Step 1: Input Validation and Linear Model Calibration
- Validates resolution profile (3 parameters for linear m/z-dependent resolution model);
- Validates retention time (RT) window consistency;
- Calculates linear resolution model parameters for m/z-dependent clustering thresholds;
- Displays reference thresholds at key m/z values (100, 200, 400, 600, 1000 Da);

#### Step 2: Data Filtering and Loading (optional RT windows)
- Selects spectra matching specified RT windows (if provided);
- Filters to MS level 1 spectra only to avoid fragmentation data;
- Separates data by ion polarity (+1 for positive, -1 for negative mode) for independent processing;
- Tracking: `pos_count` and `neg_count` monitor polarity distribution;
 
#### Step 3: Adaptive Noise Estimation and Denoising
**Purpose**: Removal of chemical noise and low-intensity artifacts while preserving true signals.

- Calculates intensity statistics from non-zero traces for unbiased noise level estimation;
- Computes adaptive noise parameters (quantile binning based on intensity distribution);
- Uses intensity-dependent noise thresholds: higher intensities use higher thresholds;
- Bins data adaptively based on signal/noise ratio and intensity distribution;
- Removes traces below the noise threshold for each intensity bin;

#### Step 4: m/z-Based Clustering
**Purpose**: Grouping of traces originating from the same molecular traces (i.e., *m/z*) across the retention time dimension. This approach accounts for instrument resolution variation across m/z range, improving clustering accuracy at both low and high m/z values.

- Clusters traces using linear m/z-dependent threshold (slope × m/z + intercept);
- Merges multiple traces within the same cluster and RT with identical intensity (keeping maximum);
- Filters clusters by minimum trace count (`minTraces`) and minimum signal-to-noise ratio (`minSNR`);

#### Step 5: Baseline Estimation and Smoothing
**Purpose**: Establishing a local intensity baseline to differentiate true peaks from background noise and smoothing intensity profiles for better derivative calculation.

- Calculates local minimum baseline using sliding window (size adapted to RT cycle time);
- Applies Savitzky-Golay filtering (window size 4, polynomial order 2) for gentle smoothing;

#### Step 6: Derivative Calculation
**Purpose**: Peak detection based on changes in intensity slope (i.e., first derivative) and curvature (i.e., second derivative).

- Computes first and second derivatives for peak localization;
- Identifies zero-crossings in the first derivative (transition from positive to negative slope);
- Each zero-crossing indicates a local maximum in the intensity profile;

#### Step 8: Peak Candidate Validation
- **Edge proximity check**: Requires minimum 2 points from cluster edges to avoid boundary artifacts;
- **Derivative validation**:
  - Pre-peak derivative (ascending slope): Averages first derivative from cluster edge until below 50% peak intensity;
  - Post-peak derivative (descending slope): Averages first derivative from peak apex until below 50% peak intensity;
  - **Dynamic window advantage**: Adapts to peak width automatically, preventing false rejections from nearby smaller peaks;
- **Local maximum check**: Ensures candidate is the highest point in its intensity range (no higher peaks on either side);
- **Criteria**: All checks must pass for acceptance (positive pre-slope, negative post-slope, confirmed local maximum);

#### Step 9: Peak Boundary Detection
- Identifies left boundary: walks backward from apex detecting valleys and intensity minima;
- Identifies right boundary: walks forward from apex detecting valleys and intensity minima;
- Enforces maximum peak width constraint to prevent merging of closely-eluting compounds;

#### Step 10: Feature Extraction and Reporting
- Consolidates all features from both ion modes;
- Compiles final feature table with RT, m/z, intensity, and quality metrics;
- Reports processing statistics: total features, polarity distribution, denoising efficiency;

### Key Implemention Aspects of the Algorithm

1. **Adaptive Noise Modeling**: Intensity-dependent noise thresholds prevent both false positives (noise spikes) and false negatives (low-intensity true signals);
2. **m/z-Aware Clustering**: Linear model-based clustering accounts for mass spectrometer resolution characteristics, improving accuracy across the m/z range;
3. **Dynamic Peak Window Detection**: Uses half-peak intensity thresholds instead of fixed windows, making the algorithm robust to peak width variation and nearby peak interference;
4. **Polarity-Separate Processing**: Processes positive and negative ion modes independently, avoiding cross-modal interference;
5. **Multi-Criterion Validation**: Requires derivative slopes, local maximum status, and edge proximity checks, reducing false positive rate while maintaining sensitivity;
6. **Derivative-Based Peak Localization**: First and second derivative analysis provides sub-sample peak localization precision and robustness to baseline variation;

## Example Datasets for Demonstration

```{r}
#| label: all-files

DT::datatable(
  files_merck_ex[, 1:3],
  options = simple_table_opt
)
```

### Curated Targets

```{r}
#| label: targets-merck

DT::datatable(db_merck_ex)
```

```{r}
#| label: peak-finding-targets-merck-function
#| include: false

run_peak_finding_debug_targets <- function(example_name) {
  db_merck_ex1 <- db_merck_ex[db_merck_ex$example %in% example_name, ]
  files <- files_merck_ex$file_path[files_merck_ex$example %in% example_name & !files_merck_ex$blank]
  
  for (i in seq_len(nrow(db_merck_ex1))) {
    cat(
      "Target ", i, ": ", db_merck_ex1$name[i],
      " - m/z: ", db_merck_ex1$mz[i],
      " - rt: ", db_merck_ex1$rt[i], "\n"
    )
    ps_ff <- DB_MassSpecMethod_FindFeatures_native(
      rtWindows = data.frame(rtmin = 300, rtmax = 3400),
      resolution_profile = c(25000L, 55000L, 80000L),
      noiseThreshold = 15,
      minSNR = 3,
      minTraces = 3,
      baselineWindow = 200,
      maxWidth = 250,
      base_quantile = 0.99,
      debug_mz = db_merck_ex1$mz[i]
    )
    clear_cache("all")
    prj <- file.path("dev", "merck_peak_finding", "data")
    ms <- DB_MassSpecEngine$new(
       = prj,
      files = files
    )
    ms$clear_cache()
    ms$Workflow = list(ps_ff)
    ms$run_workflow()
  }
  log_path <- "C:\\Users\\apoli\\Documents\\github\\StreamFind\\log"
  log_files <- list.files(log_path, pattern = "peak_detection", full.names = TRUE)
  folder_ex1 <- file.path(log_path, paste0("example_", example_name))
  if (!dir.exists(folder_ex1)) {
    dir.create(folder_ex1)
  }
  for (log_file in log_files) {
    file.copy(log_file, folder_ex1, overwrite = TRUE)
  }
  if (all(file.exists(log_files))) {
    cat("All log files copied successfully to", folder_ex1, "\n")
  } else {
    cat("Some log files were not copied successfully.\n")
  }
  for (log_file in log_files) {
    file.remove(log_file)
  }
  file.remove(list.files(prj, full.names = TRUE))
}
```

```{r}
#| label: run-peak-finding-debug-merck
#| include: false

#temp <- lapply(examples, run_peak_finding_debug_targets)
```

```{r}
#| label: debug-plots
#| results: hide
#| include: false

source("C:\\Users\\apoli\\Documents\\github\\StreamFind\\dev\\merck_peak_finding\\dev_log_plot.R")
log_path <- "C:\\Users\\apoli\\Documents\\github\\StreamFind\\log\\report_01"
log_example_folders <- list.dirs(log_path, recursive = FALSE, full.names = TRUE)
log_example_folders <- log_example_folders[grepl("example_", log_example_folders)]

plot_list <- list()

for (folder in log_example_folders) {
  log_files <- list.files(folder, pattern = "peak_detection", full.names = TRUE)
  example <- gsub(basename(folder), pattern = "example_", replacement = "")
  plot_list[[example]] <- list()
  for (log_file in log_files) {
    tryCatch({
      plot_temp <- plot_cluster_data(plot3D = FALSE, log_file = log_file)
      log_file_name <- basename(log_file)
      plot_list[[example]][[log_file_name]] <- plot_temp
    }, error = function(e) {
      cat("Error processing", log_file, ":", conditionMessage(e), "\n")
    })
  }
}
```

### Dataset `r examples[1]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-1
#| output: asis
#| echo: false

example <- examples[1]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

### Dataset `r examples[2]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-2
#| output: asis
#| echo: false

example <- examples[2]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

### Dataset `r examples[3]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-3
#| output: asis
#| echo: false

example <- examples[3]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

### Dataset `r examples[4]`

::: {.panel-tabset}

```{r}
#| label: plot-debug-example-4
#| output: asis
#| echo: false

example <- examples[4]
plot_list_example <- plot_list[[example]]
# Generate tabs with plotly plots using knitr::knit_child
for (i in seq_along(plot_list_example)) {
  log_file <- names(plot_list_example)[i]
  mz_value <- sub(".*peak_detection_(.*)\\.log", "\\1", log_file)
  # Generate tab header
  cat(sprintf("\n## m/z %s\n\n", mz_value))
  plot_list_temp <- plot_list_example[[i]]
  # Generate chunk with plot using knit_child
  plot_chunk <- knitr::knit_child(
    text = c(
      "```{r}",
      "#| echo: false",
      "#| fig-height: 7",
      sprintf("plot_list_temp"),
      "```"
    ),
    envir = environment(),
    quiet = TRUE
  ) 
  cat(plot_chunk, sep = "\n")
  cat("\n")
}
```

:::

## Comparison with Other Algorithms

The C++ native peak finding is compared against OpenMS and XCMS3 (centWave).

```{r}
#| label: create-engine
#| eval: false

engine <- MassSpecEngine$new(
  analyses = files_merck_ex$file_path[!files_merck_ex$blank]
)

engine
```

### Algorithms

```{r}
#| label: ps-ff-definitions
#| eval: false

ps_ff_sf <- MassSpecMethod_FindFeatures_native(
  rtWindows = data.frame(rtmin = 300, rtmax = 3400),
  resolution_profile = c(25000L, 55000L, 80000L),
  noiseThreshold = 15,
  minSNR = 3,
  minTraces = 3,
  baselineWindow = 200,
  maxWidth = 250,
  base_quantile = 0.99,
  debug_mz = 0
)

ps_ff_openms <- MassSpecMethod_FindFeatures_openms(
  noiseThrInt = 15,         # Aligned with native noiseThreshold
  chromSNR = 3,             # Aligned with native minSNR
  chromFWHM = 7,
  mzPPM = 15,
  reEstimateMTSD = TRUE,
  traceTermCriterion = "sample_rate",
  traceTermOutliers = 5,
  minSampleRate = 1,
  minTraceLength = 3,       # Aligned with native minTraces (approximate - native uses count, openms uses seconds)
  maxTraceLength = 250,     # Aligned with native maxWidth
  widthFiltering = "fixed",
  minFWHM = 4,
  maxFWHM = 250,            # Aligned with native maxWidth
  traceSNRFiltering = TRUE,
  localRTRange = 0,
  localMZRange = 0,
  isotopeFilteringModel = "none",
  MZScoring13C = FALSE,
  useSmoothedInts = FALSE,
  intSearchRTWindow = 3,
  useFFMIntensities = FALSE,
  verbose = FALSE
)

ps_ff_xcms <- MassSpecMethod_FindFeatures_xcms3_centwave(
  ppm = 15,                 # Aligned with native (uses resolution_profile, ~15 ppm)
  peakwidth = c(5, 250),    # Aligned with native maxWidth (250 seconds)
  snthresh = 3,             # Aligned with native minSNR
  prefilter = c(3, 15),     # Aligned with native minTraces (3) and noiseThreshold (15)
  mzCenterFun = "wMean",
  integrate = 1,
  mzdiff = -2e-04,
  fitgauss = TRUE,
  noise = 15,               # Aligned with native noiseThreshold
  verboseColumns = TRUE,
  firstBaselineCheck = FALSE,
  extendLengthMSW = FALSE
)
```

```{r}
#| label: run-speed-test
#| include: false

#source("C:\\Users\\apoli\\Documents\\github\\StreamFind\\dev\\merck_peak_finding\\run_speed_test.R")
results_dir <- "C:\\Users\\apoli\\Documents\\github\\StreamFind\\dev\\merck_peak_finding"
all_results <- fread(file.path(results_dir, "speed_test_all_results.csv"))
summary_stats <- fread(file.path(results_dir, "speed_test_summary_stats.csv"))
all_intensity_data <- fread(file.path(results_dir, "speed_test_intensity_data.csv"))
intensity_summary <- fread(file.path(results_dir, "speed_test_intensity_summary.csv"))
```

### Speed Test Summary

```{r}
#| label: display-speed-test-summary

# Display summary statistics
DT::datatable(
  summary_stats,
  options = list(
    dom = 't',
    paging = FALSE
  ),
  caption = "Speed Test - Summary Statistics"
) %>%
  DT::formatRound(columns = c("mean_time", "sd_time", "features_per_second"), digits = 2) %>%
  DT::formatRound(columns = c("mean_features"), digits = 0)
```

### Intensity Distribution Comparison

```{r}
#| label: intensity-distribution-histograms
#| results: hide

# Create interactive plotly histogram data for comparison
if (!is.null(all_intensity_data) && nrow(all_intensity_data) > 0) {
  # Calculate histogram bins for each algorithm
  n_bins <- 50
  
  # Create color palette
  colors <- c("#F8766D", "#00BA38", "#619CFF")
  algorithms <- unique(all_intensity_data$algorithm)
  
  # Histogram on log10 scale (Plotly)
  p1 <- plot_ly() %>%
    layout(
      title = "Feature Intensity Distribution (Log10 Scale)",
      xaxis = list(title = "Log10(Intensity + 1)"),
      yaxis = list(title = "Count"),
      barmode = "overlay",
      legend = list(x = 0.5, y = -0.2, orientation = "h", xanchor = "center")
    )
  
  for (i in seq_along(algorithms)) {
    alg <- algorithms[i]
    data_subset <- all_intensity_data[algorithm == alg, log10_intensity]
    p1 <- p1 %>%
      add_histogram(
        x = data_subset,
        name = alg,
        opacity = 0.6,
        marker = list(color = colors[i]),
        nbinsx = n_bins
      )
  }
  
  # Density plot on log10 scale (Plotly)
  p2 <- plot_ly() %>%
    layout(
      title = "Feature Intensity Density (Log10 Scale)",
      xaxis = list(title = "Log10(Intensity + 1)"),
      yaxis = list(title = "Density"),
      legend = list(x = 0.5, y = -0.2, orientation = "h", xanchor = "center")
    )
  
  for (i in seq_along(algorithms)) {
    alg <- algorithms[i]
    data_subset <- all_intensity_data[algorithm == alg, log10_intensity]
    density_result <- density(data_subset, n = 512)
    p2 <- p2 %>%
      add_lines(
        x = density_result$x,
        y = density_result$y,
        name = alg,
        fill = "tozeroy",
        fillcolor = paste0(colors[i], "4D"),
        line = list(color = colors[i], width = 2)
      )
  }
  
  # Box plot comparison - Log10 scale (Plotly)
  p3_log <- plot_ly() %>%
    layout(
      title = "Feature Intensity Distribution by Algorithm (Log10 Scale)",
      xaxis = list(title = "Algorithm"),
      yaxis = list(title = "Log10(Intensity + 1)"),
      legend = list(x = 0.5, y = -0.2, orientation = "h", xanchor = "center")
    )
  
  for (i in seq_along(algorithms)) {
    alg <- algorithms[i]
    data_subset <- all_intensity_data[algorithm == alg, log10_intensity]
    p3_log <- p3_log %>%
      add_boxplot(
        y = data_subset,
        name = alg,
        fillcolor = paste0(colors[i], "4D"),
        marker = list(
          color = colors[i],
          outliercolor = colors[i]
        ),
        line = list(color = colors[i]),
        boxmean = TRUE
      )
  }
  
  # Box plot comparison - Actual intensity scale (Plotly)
  p3_actual <- plot_ly() %>%
    layout(
      title = "Feature Intensity Distribution by Algorithm",
      xaxis = list(title = "Algorithm"),
      yaxis = list(title = "Intensity"),
      legend = list(x = 0.5, y = -0.2, orientation = "h", xanchor = "center")
    )
  
  for (i in seq_along(algorithms)) {
    alg <- algorithms[i]
    data_subset <- all_intensity_data[algorithm == alg, intensity]
    p3_actual <- p3_actual %>%
      add_boxplot(
        y = data_subset,
        name = alg,
        fillcolor = paste0(colors[i], "4D"),
        marker = list(
          color = colors[i],
          outliercolor = colors[i]
        ),
        line = list(color = colors[i]),
        boxmean = TRUE
      )
  }
  
} else {
  cat("No intensity data available for comparison.\n")
}
```

```{r}
#| label: plot-intensity-distribution-histograms
#| fig-width: 12
#| fig-height: 8

p1
```

```{r}
#| label: plot-intensity-distribution-density
#| fig-width: 12
#| fig-height: 8

p2
```

```{r}
#| label: plot-intensity-distribution-boxplot-log
#| fig-width: 12
#| fig-height: 8

p3_log
```

```{r}
#| label: plot-intensity-distribution-boxplot-actual
#| fig-width: 12
#| fig-height: 8

p3_actual
```

```{r}
#| label: display-intensity-summary

# Display intensity distribution summary table
if (!is.null(intensity_summary) && nrow(intensity_summary) > 0) {
  DT::datatable(
    intensity_summary,
    options = list(
      dom = 't',
      paging = FALSE
    ),
    caption = "Intensity Distribution Summary by Algorithm"
  ) %>%
    DT::formatRound(columns = c("mean_intensity", "median_intensity", "sd_intensity", 
                                  "q25_intensity", "q75_intensity", "min_intensity", 
                                  "max_intensity"), digits = 0)
} else {
  cat("No intensity summary data available.\n")
}
```

## Preliminary StreamFind Workflow

```{r}
#| label: create-demo-engine

demo <- DB_MassSpecEngine$new(
   = file.path("data"),
  files = files_merck_ex$file_path[c(1, 2)]
)

set_replicate_names(demo$Analyses, c("blank", "sample"))
set_blank_names(demo$Analyses, rep("blank", 2))

demo
```

### Inspect TIC

```{r}
#| label: plot-demo-tic

plot_spectra_tic(demo$Analyses, downsize = 5, levels = 1)
```

### Workflow

```{r}
#| label: run-demo-workflow

workflow <- list(
  DB_MassSpecMethod_FindFeatures_native(
    rtWindows = data.frame(rtmin = 300, rtmax = 3400),
    resolution_profile = c(25000L, 55000L, 80000L),
    noiseThreshold = 15,
    minSNR = 3,
    minTraces = 3,
    baselineWindow = 200,
    maxWidth = 250,
    base_quantile = 0.99,
    debug_mz = 0
  )
)

demo$Workflow <- workflow

demo
```

```{r}
#| label: run-demo-workflow-execute

demo$run_workflow()
```

### Results

```{r}
#| label: show-demo-results
#| fig-height: 8

plot_features(
  demo$NonTargetAnalysis,
  analyses = 2,
  mz = db_merck_ex[db_merck_ex$example %in% examples[1], ],
  ppm = 20,
  sec = 60
)
```

```{r}
#| label: clean-demo-data
#| include: false

file.remove(list.files("data", full.names = TRUE))
```
