---
title: 'Quantification of Intact Monoclonal Antibody with LC-DAD'
author:
  name: Ricardo Cunha
  email: cunha@iuta.de
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    fig_caption: true
    toc: true
    number_sections: true
    toc_float: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{'Quantification of Intact Monoclonal Antibody with LC-DAD'}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(plotly)
library(data.table)
library(StreamFind)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 9,
  results = "markup",
  comment = ""
)

options(
  DT.options = list(
    dom = "lfrtip",
    autoWidth = TRUE,
    pageLength = 10,
    scrollX = TRUE,
    lengthMenu = c(10, 20, 50, 75, 100),
    initComplete = htmlwidgets::JS(
      "function(settings, json) {",
      "$(this.api().table().body()).css({'font-size': '12px'});",
      "$(this.api().table().body()).css({'white-space': 'nowrap'});",
      "}"
    )
  )
)

#path <- "E:/example_files/isoft/240819_BVCZ"
path <- "C:/Users/apoli/Documents/example_files/isoft/240819_BVCZ"
files <- list.files(path = path, pattern = "\\.d$", full.names = TRUE)[-15]
```

<br>
<br>

***

# Introduction

This document demonstrates the quantification of the intact monoclonal antibody Bevacizumab using Liquid Chromatography with Diode Array Detection (LC-DAD). The files include a blank, a calibration set with 10 points from 0.5 to 5 mg/L, a QC triplicate with 2.5 mg/L, and a sample in triplicate to be quantified. The data was acquired with an Agilent 1260 Infinity II system coupled to a DAD detecter. The raw data files are in Agilent's .d format.

```{r files}
basename(files)
```

# Processing Engine

The `StreamFind` package provides a flexible and powerful framework for processing mass spectrometry data. In this example, we will use the `MassSpecEngine` class to handle the LC-DAD data. The engine will be initialized with the raw files, and we will specify that the data should be centroided and processed at level 1. Note that The `msconvert` from [ProteoWizard](https://proteowizard.sourceforge.io/) was used in the background to convert the files to the open format mzML.

```{r engine}
# Starts engine for quantification
ms <- StreamFind::MassSpecEngine$new(
  metadata = list(name = "Quantification of Bevacizumab"),
  analyses = files,
  centroid = TRUE,
  levels = 1
)

# Edit to assign analysis replicate groups
ms$Analyses <- set_replicate_names(
  ms$Analyses,
  c(
    "cal_0.5",
    "cal_1",
    "cal_1.5",
    "cal_2",
    "cal_2.5",
    rep("Sample", 3),
    "cal_3",
    "cal_3.5",
    "cal_4",
    "cal_4.5",
    "cal_5",
    "Blank",
    rep("QC_2.5", 3)
  )
)

# Concentration was obtained by attempting to convert the file names to numeric
# Alternatively, you can use the setter method ms$Analyses$concentrations to manually add the
# concentration values with NA_Real_ when the concentration is to be calculated
DT::datatable(
  info(ms$Analyses)
)
```

# Available Chromatograms

```{r available-chroms}
chroms = data.frame(
  "name" = unique(ms$Analyses$analyses[[1]]$chromatograms_headers$id)
)

DT::datatable(chroms)
```

The chromatogram of interest is the `r data.frame("name" = unique(ms$Analyses$analyses[[1]]$chromatograms_headers$id))$name[2]`, which is the second, and we know that our analyte elutes between 250 and 400 seconds.

# Workflow

```{r load-chroms}
ms$run(
  MassSpecMethod_LoadChromatograms_native(
    chromatograms = chroms$name[2],
    rtmin = 250,
    rtmax = 400
  )
)
```

```{r plot-chroms}
plot_chromatograms(
  ms$Results$MassSpecResults_Chromatograms,
  colorBy = "replicates",
  normalized = FALSE
)
```

```{r correct-chroms}
ms$run(
  MassSpecMethod_SmoothChromatograms_movingaverage(
    windowSize = 3
  )
)

ms$run(
  MassSpecMethod_CorrectChromatogramsBaseline_airpls(
    lambda = 25,
    differences = 1,
    itermax = 20
  )
)
```

```{r plot-chroms-corrected}
plot_chromatograms(
  ms$Results$MassSpecResults_Chromatograms,
  colorBy = "replicates",
  normalized = FALSE
)
```

```{r integrate-chroms}
ms$run(
  MassSpecMethod_IntegrateChromatograms_pracma(
    merge = TRUE,
    closeByThreshold = 5,
    minPeakHeight = 10,
    minPeakDistance = 3,
    minPeakWidth = 5,
    maxPeakWidth = 50,
    minSN = 1
  )
)
```

```{r plot-chroms-integrated}
plot_chromatograms_peaks(
  ms$Results$MassSpecResults_Chromatograms,
  colorBy = "replicates+targets"
)
```

```{r quantify-chroms}
# Note that concentration values of the calibration curve
# were automatically obtained by the file name
ms$run(
  MassSpecMethod_QuantifyChromatographicPeaks_native(
    calibration = NA_real_,
    value = "area",
    model = "linear"
  )
)
```

```{r plot-calibration}
model <- ms$Results$MassSpecResults_Chromatograms$calibration_model

fitted_values <- stats::predict(model)
r_squared <- summary(model)$r.squared

dt_cal <- data.table::data.table(
  fitted_values = fitted_values,
  calibration_values = model$model$calibration,
  values = model$model$values
)

fig <- plotly::plot_ly(
  model$model,
  x = ~calibration_values,
  y = ~values,
  type = 'scatter',
  mode = 'markers',
  name = "Data"
) %>%
  add_lines(x = ~fitted_values, y = ~values, name = "Fitted Line") %>%
  layout(
    title = paste("Linear Regression (R² =", round(r_squared, 4), ")"),
    xaxis = list(title = "mg/L"),
    yaxis = list(title = "Intensity / counts"),
    showlegend = TRUE
  )

fig
```

```{r quantification-results}
peaks <- get_chromatograms_peaks(ms$Results$MassSpecResults_Chromatograms)
DT::datatable(peaks)
```
