---
title: 'Wastewater ozonation showcase'
subtitle: |
  | Mass spectrometry project example
  | <br>
author:
  name: Ricardo Cunha
  email: cunha@iuta.de
  # affiliation: "Institut für Umwelt & Energie, Technik & Analytik e. V. (IUTA) <br> Bliersheimer Str. 58 - 60, 47229 Duisburg"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    fig_caption: true
    toc: true
    number_sections: true
    toc_float: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Developer guide for mass spectrometry}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Libraries
library(knitr)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(plotly)
requireNamespace("xcms", quietly = TRUE)
library(StreamFind)

# Variables
logo <- NA_character_

# Global options
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width = 9, results = "asis")

# Working directory
path_wd <- getwd()

# Add logo file
path_extdata <- system.file(package = "StreamFindData", dir = "extdata")
logo <- file.path(path_extdata, "logo_StreamFind.png")
```

```{r logo, echo=FALSE}
# if (!is.na(logo)) {
#   htmltools::img(
#   src = knitr::image_uri(logo),
#   alt = "logo",
#   style =
#     paste0("
#       margin-left:inherit;
#       position:absolute;
#       top:150px;
#       left:1200px;
#       width:", 250, "px;
#       height:", 250*(1633/3070), "px;"
#     )
#   )
# }
```

```{r resources, include=FALSE}
# MS files
all_files <- StreamFindData::get_all_file_paths()
files <- all_files[grepl("blank|influent|o3sw", all_files)]

# Chemicals database
db_all <- StreamFindData::get_tof_spiked_chemicals()
db_all <- db_all[grepl("S", db_all$tag), ]
cols <- c("name", "formula", "mass", "rt", "tag")
db_all <- db_all[, cols, with = FALSE]

db_is <- db_all[db_all$tag %in% "IS", ]
db_is <- db_is[!db_is$name %in% c("Ibuprofen-d3", "Naproxen-d3"), ]
db <- db_all[db_all$tag %in% "S", ]
```

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

<br>

<br>

***

# Introduction

In this article, we demonstrate how the StreamFind R package can be used to evaluate ozonation of secondary wastewater effluent (i.e., effluent of the aerated biological treatment) using mass spectrometry (MS). A set of 18 *mzML* files, representing blank, influent and effluent measurements in triplicate for both positive and negative ionization mode, are used.

```{r files, results='markup'}
basename(files)
```

The showcase will use the StreamFind main class *MassSpecData*, which encapsulates all tools required for parsing, storing, processing and visualizing MS data. Note that not all methods/functions will be shown as the demonstration focuses of the workflow to assess wastewater ozonation. For a complete overview of the StreamFind R package for mass spectrometry, the article [StreamFind for mass spectrometry](XXX) is recommended.

# *MassSpecData*

The main [*R6*](https://r6.r-lib.org/) [*MassSpecData*](https://odea-project.github.io/StreamFind/reference/MassSpecData.html) class object is created using the `MassSpecData$new()`, as shown below. The argument `files` can be used to add directly the set of *mzML* files. Note that the original vendor files were converted to *mzML* format using the [convert_ms_files()](https://odea-project.github.io/StreamFind/reference/convert_ms_files.html) function, which uses the *msConvert* command line from [ProteoWizard](https://proteowizard.sourceforge.io/download.html).

```{r create-ms, message=TRUE, warning=FALSE}
# Create a MassSpecData from mzML files
ms <- MassSpecData$new(files = files)
```

```{r show-ms, results='markup'}
# Print in console a summary of the MassSpecData
ms
```

# *ProjectHeaders*

Project headers (e.g., name, author and description) can be added to the *MassSpecData* using the `MassSpecData$add_headers()`. The headers are converted to an S3 [*ProjectHeaders*](https://odea-project.github.io/StreamFind/reference/ProjectHeaders.html) class object in the *MassSpecData*, where it goes through a validation to ensure conformity (more details [here](https://odea-project.github.io/StreamFind/reference/ProjectHeaders.html)).

```{r add-headers, results='markup'}
# Add headers to the MassSpecData
ms$add_headers(
  name = "Wastewater Ozonation Showcase",
  author = "Ricardo Cunha",
  description = "Demonstration project"
)

# Get the headers as ProjectHeaders
ms$get_headers()

# Get the MassSpecData date
ms$get_headers(value = "date")
```

# Replicates and blanks

The analysis replicate names and the associated blank replicate name can be amended in the *MassSpecData*, as shown below. Alternatively, a `data.frame` with column names *file*, *replicate* and *blank* could be added as the `files` argument in `MassSpecData$new(files = files)` to have directly the replicate and blank replicate names assigned (more details [here](https://odea-project.github.io/StreamFind/reference/MassSpecData.html#method-MassSpecData-new)).

```{r add-replicates, results='markup'}
# Character vector with analysis replicate names
rpls <- c(
  rep("blank_neg", 3),
  rep("blank_pos", 3),
  rep("in_neg", 3),
  rep("in_pos", 3),
  rep("out_neg", 3),
  rep("out_pos", 3)
)

# Character vector with associated blank replicate names
# Note that the order should match the respective replicate
blks <- c(
  rep("blank_neg", 3),
  rep("blank_pos", 3),
  rep("blank_neg", 3),
  rep("blank_pos", 3),
  rep("blank_neg", 3),
  rep("blank_pos", 3)
)

# Chaining add replicates and blanks
ms$add_replicate_names(rpls)$add_blank_names(blks)

# Replicates and blanks were amended
ms$get_overview()[, 1:5]
```

# *ProcessingSettings*

Data processing is performed by modules according to [*ProcessingSettings*](https://odea-project.github.io/StreamFind/reference/ProcessingSettings.html). The S3 *ProcessingSettings* class objects are obtained via the respective [`Settings_[module name]_[algorithm name]`](https://odea-project.github.io/StreamFind/reference/index.html#processing-settings-for-ms-data) constructor functions, attributing the respective subclass. Below we obtain the *ProcessingSettings* for the module `find_features()` using the algorithm *xcms3_centwave*. The parameters for each processing module can be changed via the constructor arguments. Documentation for each *ProcessingSettings* subclass can be found in the [StreamFind reference documentation](https://odea-project.github.io/StreamFind/reference/index.html).

```{r ffs, results='markup'}
# Get ProcessingSettings for finding features using the xcms3 centwave algorithm
ffs <- Settings_find_features_xcms3_centwave()

# Print in console the details of the ProcessingSettings
ffs
```

```{r other-settings, results='markup'}
# Modifying the parameters with the constructor arguments
fls <- Settings_filter_features_StreamFind(
  minIntensity = 5000,
  minSnRatio = 20,
  maxGroupSd = 30,
  blank = 5,
  minGroupAbundance = 3,
  excludeIsotopes = TRUE
)

fls

afs <- Settings_annotate_features_StreamFind()

afs

gfs <- Settings_group_features_xcms3_peakdensity()

gfs
```

When all required *ProcessingSettings* are created, they can be added or used directly to run the workflow. In the example below, the *ProcessingSettings* are first added to the *MassSpecData* in the desired order.

```{r add-settings, results='markup'}
# Add settings for finding, annotating, grouping and finally filter features
ms$add_settings(settings = list(ffs, afs, gfs, fls))

# Getting the workflow settings
ms$get_settings_names()
```

# `run_workflow()`

The modules assigned by the added *ProcessingSettings* can be processed by `run_workflow()`, as demonstrated below. Note that with `run_workflow()`, the processing modules are applied with the same order as they were added.

```{r do-workflow, results='markup', warning=FALSE}
# Run all ProcessingSettings added to the MassSpecData
ms$run_workflow()
```

Alternatively, all modules can be processed by chaining, as shown below. By chaining, the order can be modified as the modules are individually called.

```{r do-workflow-chain, eval=FALSE}
# Alternatively, the modules can be chained
ms$find_features()$annotate_features()$group_features()$filter_features()
```

# Results

The created features and feature groups can be inspected as `data.table` objects or plotted by dedicated modules in the *MassSpecData*.

## `data.table` objects

The features and feature groups can be obtained as `data.table` with the `MassSpecData$get_features()` and `MassSpecData$get_groups()` methods. The methods also allow to look for specific features/feature groups using mass, mass-to-charge ratio, retention time and drift time targets, as show below for a small set of compound targets where mass and retention time expected value are known. Note that drift time is only applicable for MS data with ion mobility separation.

```{r database-show, results='markup'}
db
```

```{r show-target-groups, results='markup'}
# compounds are searched by monoisotopic mass and retention time
# ppm and sec set the mass (im ppm) and time (in seconds) allowed deviation, respectively
# average applied a mean to the intensities in each analysis replicate group
ms$get_groups(mass = db, ppm = 5, sec = 10, average = TRUE)
```

Already by inspection of the `data.table`, it is possible to see compounds detected in the influent but not in the effluent (e.g., Carbamazepine) or compounds that are appear to be reduced during ozonation (e.g., Metoprolol). Since positive and negative ionization mode were combined, there are compounds that appear in both polarities and are grouped by neutral monoisotopic mass (e.g., Diclofenac).

## `plot_groups` methods

For a better overview of the results, the method `MassSpecData$plot_groups()` or even more detailed the method `MassSpecData$plot_groups()` can be used.

```{r plot-groups}
# set legendNames to TRUE for using the names in db as legend
ms$plot_groups(mass = db, ppm = 5, sec = 10, legendNames = TRUE)
```

```{r plot-groups-overview, fig.height=10}
ms$plot_groups_overview(mass = db, ppm = 5, sec = 10, legendNames = TRUE)
```

## Filtered not removed

The `filter_features()` module was applied to filter features out according to defined conditions/thresholds. The filtered features were not removed but just tagged as filtered and a filter tag was assigned. For instance, when the method `MassSpecData$get_features()` is run with `filtered` argument set to `TRUE`, the filtered features are also shown. Below, we search for the internal standards added to each analysis that were filtered when applying the blank subtraction, receiving the tag **blank** in the *filter* column and `TRUE` in the *filtered* column.

```{r database-is-show, results='markup'}
db_is
```

```{r show-filtered-is, results='markup'}
# set filtered to TRUE for showing filtered features/feature groups
ms$get_groups(mass = db_is, ppm = 8, sec = 10, average = TRUE, filtered = TRUE)
```

## Isotopes

The method `annotate_features()` was applied to annotate the natural isotopes within the MS data. Implementation of adducts and in source fragments annotation is planned but not yet available with the StreamFind algorithm for annotation of features (i.e., the object `afs`). The method `MassSpecData$get_components()` can be used to search for components, as shown below for analyses number 11. Because the filter `excludeIsotopes` was applied, the isotopic features are filtered out with tag **isotope** or **minIntensity** when filtered by `minIntensity`, which was applied first. Although the first isotope of Diclofenac was not found, the second was annotated as the `maxGaps` argument in `afs` was set to 1. This is a frequent case for compounds with halogens.

```{r show-isotopes, results='markup'}
# Components of Diclofenac and Candesartan in analysis 11
ms$get_components(
  analyses = 11,
  mass = db[db$name %in% c("Diclofenac", "Candesartan"), ],
  ppm = 5, sec = 10
)
```

The components can also be visualized with the method `MassSpecData$map_components()`, as shown below for the internal standards added to analysis 11. Note that again the `filtered` argument was set to `TRUE` to return also filtered features.

```{r plot-isotopes, results='markup'}
ms$map_components(
  analyses = 11,
  mass = db_is,
  ppm = 8, sec = 10,
  filtered = TRUE,
  legendNames = TRUE
)
```

# More to come

Future implementations to be included in this showcase are:

- Correcting matrix for better comparison between influent and effluent
- Annotation of adducts and in-source fragments
- Suspect screening with MS<sup>2</sup> validation and formula/compound assignment via patRoon
- Screening of transformation products using the *biotransformer* tool via patRoon

***

<!-- <style> -->
<!-- body, p { -->
<!--   color: black; -->
<!--   font-family: Arial; -->
<!--   text-align: justify; -->
<!--   font-size: 12pt; -->
<!-- } -->
<!-- h1{ -->
<!--   font-size: 20pt; -->
<!--   font-style: bold; -->
<!-- } -->
<!-- h2{ -->
<!--   font-size:18pt; -->
<!--   font-style: bold; -->
<!-- } -->
<!-- h3{ -->
<!--   font-size:14pt; -->
<!--   font-style: bold; -->
<!-- } -->
<!-- .list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover { -->
<!--     background-color: #177013; -->
<!-- } -->
<!-- caption { -->
<!--   color: black; -->
<!--   font-size: 1.0em; -->
<!-- } -->
<!-- </style> -->
