---
title: 'Non-target Analysis for Wastewater Treatment Evaluation'
author:
  name: Ricardo Cunha
  email: cunha@iuta.de
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    fig_caption: true
    toc: true
    number_sections: true
    toc_float: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{'Non-target Analysis for Wastewater Treatment Evaluation'}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Libraries
library(knitr)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(plotly)
library(StreamFind)

knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.width = 9,
  results = "markup",
  comment = "",
  message = FALSE,
  warning = FALSE
)
```

```{r resources, include=FALSE}
all_files <- StreamFindData::get_ms_file_paths()
files <- all_files[grepl("blank|influent|o3sw", all_files)]
db_all <- StreamFindData::get_ms_tof_spiked_chemicals()
db_all <- db_all[grepl("S", db_all$tag), ]
cols <- c("name", "formula", "mass", "rt", "tag")
db_is <- db_all[db_all$tag %in% "IS", ]
db_is <- db_is[, cols, with = FALSE]
db_is <- db_is[!db_is$name %in% c("Ibuprofen-d3", "Naproxen-d3"), ]
db <- db_all[db_all$tag %in% "S", ]
db <- db[, cols, with = FALSE]
db_with_ms2 <- StreamFindData::get_ms_tof_spiked_chemicals_with_ms2()
db_with_ms2 <- db_with_ms2[db_with_ms2$tag %in% "S", ]
db_with_ms2 <- db_with_ms2[, c("name", "formula", "mass", "SMILES", "rt", "polarity", "fragments"), with = FALSE]
db_with_ms2$polarity[db_with_ms2$polarity == 1] <- "positive"
db_with_ms2$polarity[is.na(db_with_ms2$polarity)] <- "positive"
db_with_ms2$polarity[db_with_ms2$polarity == -1] <- "negative"
```

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

<br>

<br>

***

# Introduction

In this article we demonstrate how StreamFind can be used to evaluate ozonation of secondary wastewater effluent (i.e., effluent of the aerated biological treatment) using mass spectrometry (MS). A set of 18 *mzML* files are used, representing blank, influent and effluent measurements in triplicate for both positive and negative ionization modes.

```{r files}
basename(files)
```

The showcase will use the StreamFind *MassSpecEngine*, which encapsulates all tools required for parsing, storing, processing and visualizing MS data. Note that not all methods/functions will be shown, as the demonstration focuses of the workflow to assess wastewater treatment. Other processing methods for MS data are available in the StreamFind package and can be found in the [StreamFind reference documentation](https://odea-project.github.io/StreamFind/reference/index.html).

# *MassSpecEngine*

The [*R6*](https://r6.r-lib.org/) [*MassSpecEngine*](https://odea-project.github.io/StreamFind/reference/MassSpecEngine.html) class object is created using `MassSpecEngine$new()`, as shown below. The `analyses` argument can be used to add the set of ms files directly. In this demonstration, we use *mzML* files. However, the original ms vendor files can also be used, but they will be converted to *mzML* format using *msConvert* from [ProteoWizard](https://proteowizard.sourceforge.io/download.html) in the background.

```{r create-ms}
# Creates a MassSpecEngine from mzML files
ms <- MassSpecEngine$new(
  metadata = list(name = "Wastewater NTA"),
  analyses = files
)
```

```{r show-ms}
# Engine class hierarchy
class(ms)

# Analyses class hierarchy
class(ms$Analyses)
```

# *Metadata*

Project metadata (e.g., name, author and description) can be added to the `MassSpecEngine$Metadata` as a named list object as shown below. The elements of the list can be anything but must have length one.

```{r add-headers}
# Adds metadata to the MassSpecEngine
ms$Metadata <- list(
  name = "Wastewater Ozonation Showcase",
  author = "Ricardo Cunha",
  description = "Demonstration project"
)

# Gets and shows the metadata
show(ms$Metadata)

# Gets the date element from the Metadata
ms$Metadata$date
```

# Replicates and blanks

The analysis replicate names and the associated blank replicate name can be amended in the *MassSpecEngine*, as shown below. Alternatively, a `data.frame` with column names *file*, *replicate* and *blank* could be added as the `analyses` argument in `MassSpecEngine$new(analyses = files)` to have directly the replicate and blank replicate names assigned (more details [here](https://odea-project.github.io/StreamFind/reference/MassSpecEngine.html#method-MassSpecEngine-new)).

```{r add-replicates}
# Character vector with analysis replicate names
rpls <- c(
  rep("blank_neg", 3),
  rep("blank_pos", 3),
  rep("influent_neg", 3),
  rep("influent_pos", 3),
  rep("effluent_neg", 3),
  rep("effluent_pos", 3)
)

# Character vector with associated blank replicate names
# Note that the order should match the respective replicate
blks <- c(
  rep("blank_neg", 3),
  rep("blank_pos", 3),
  rep("blank_neg", 3),
  rep("blank_pos", 3),
  rep("blank_neg", 3),
  rep("blank_pos", 3)
)

# Amends replicate and blank names
ms$Analyses <- set_replicate_names(ms$Analyses, rpls)
ms$Analyses <- set_blank_names(ms$Analyses, blks)

# Replicates and blanks were amended
info(ms$Analyses)[, c(1:3, 5)]
```

# *ProcessingStep*

Data processing is performed by steps according to *ProcessingStep* objects. *ProcessingStep* class objects are obtained via the respective [`[Engine type]Method_[method name]_[algorithm name]`](https://odea-project.github.io/StreamFind/reference/index.html#processing-settings-for-ms-data) constructor functions, attributing the respective subclass. Below we obtain the *ProcessingStep* for the method `FindFeatures` using the algorithm *openms*. The parameters can be changed via the constructor arguments. Documentation for each *ProcessingStep* subclass can be found in the [StreamFind reference documentation](https://odea-project.github.io/StreamFind/reference/index.html).

```{r ffs}
# Gets ProcessingStep for finding features using the openms algorithm
ffs <- MassSpecMethod_FindFeatures_openms(
  noiseThrInt = 1000,
  chromSNR = 3,
  chromFWHM = 7,
  mzPPM = 15,
  reEstimateMTSD = TRUE,
  traceTermCriterion = "sample_rate",
  traceTermOutliers = 5,
  minSampleRate = 1,
  minTraceLength = 4,
  maxTraceLength = 70,
  widthFiltering = "fixed",
  minFWHM = 4,
  maxFWHM = 35,
  traceSNRFiltering = TRUE,
  localRTRange = 0,
  localMZRange = 0,
  isotopeFilteringModel = "none",
  MZScoring13C = FALSE,
  useSmoothedInts = FALSE,
  intSearchRTWindow = 3,
  useFFMIntensities = FALSE,
  verbose = FALSE
)

# Prints in console the details of the ProcessingStep
show(ffs)
```

```{r other-methods}
# Creates an ordered list with all processing steps for the MS data
workflow <- list(

  # Find features using the openms algorithm, created above
  ffs,

  # Annotation of natural isotopes and adducts
  MassSpecMethod_AnnotateFeatures_StreamFind(
    rtWindowAlignment = 0.3,
    maxIsotopes = 8,
    maxCharge = 2,
    maxGaps = 1
  ),

  # Excludes annotated isotopes and adducts
  MassSpecMethod_FilterFeatures_StreamFind(
    excludeIsotopes = TRUE,
    excludeAdducts = TRUE
  ),

  # Grouping features across analyses
  MassSpecMethod_GroupFeatures_openms(
    rtalign = FALSE,
    QT = FALSE,
    maxAlignRT = 5,
    maxAlignMZ = 0.008,
    maxGroupRT = 5,
    maxGroupMZ = 0.008,
    verbose = FALSE
  ),

  # Filter feature groups with maximum intensity below 5000 counts
  MassSpecMethod_FilterFeatures_StreamFind(
    minIntensity = 3000
  ),

  # Fill features with missing data
  # Reduces false negatives
  MassSpecMethod_FillFeatures_StreamFind(
    withinReplicate = FALSE,
    rtExpand = 2,
    mzExpand = 0.0005,
    minTracesIntensity = 1000,
    minNumberTraces = 6,
    baseCut = 0.3,
    minSignalToNoiseRatio = 3,
    minGaussianFit = 0.2
  ),

  # Calculate quality metrics for each feature
  MassSpecMethod_CalculateFeaturesQuality_StreamFind(
    filtered = FALSE,
    rtExpand = 2,
    mzExpand = 0.0005,
    minTracesIntensity = 1000,
    minNumberTraces = 6,
    baseCut = 0
  ),

  # Filter features based on minimum signal-to-noise ratio (s/n)
  # The s/n is calculated using the CalculateFeaturesQuality method
  MassSpecMethod_FilterFeatures_StreamFind(
    minSnRatio = 5
  ),

  # Filter features using other parameters via the patRoon package
  MassSpecMethod_FilterFeatures_patRoon(
    maxReplicateIntRSD = 40,
    blankThreshold = 5,
    absMinReplicateAbundance = 3
  ),

  # Finds internal standards in the MS data
  # db_is is a data.table with the
  # name, mass and expected retention time of
  # spiked internal standards, as shown below
  MassSpecMethod_FindInternalStandards_StreamFind(
    database = db_is,
    ppm = 8,
    sec = 10
  ),
  
  # Corrects matrix suppression using the TiChri method from
  # 10.1021/acs.analchem.1c00357 to better compare influent and effluent
  MassSpecMethod_CorrectMatrixSuppression_TiChri(
    mpRtWindow = 10,
    istdAssignment = "range",
    istdRtWindow = 50,
    istdN = 2
  ),

  # Loads MS1 for features not filtered
  MassSpecMethod_LoadFeaturesMS1_StreamFind(
    filtered = FALSE
  ),

  # Loads MS2 for features not filtered
  MassSpecMethod_LoadFeaturesMS2_StreamFind(
    filtered = FALSE
  ),

  # Loads feature extracted ion chromatograms (EIC)
  MassSpecMethod_LoadFeaturesEIC_StreamFind(
    filtered = FALSE
  ),

  # Performs suspect screening using the StreamFind algorithm
  # db_with_ms2 is a database with suspect chemical standards
  # includes MS2 data (i.e., fragmentation pattern) from standards
  MassSpecMethod_SuspectScreening_StreamFind(
    database = db_with_ms2,
    ppm = 10,
    sec = 15,
    ppmMS2 = 10,
    minFragments = 3
  )
)
```

Then, the list can be added to the *MassSpecEngine*. Note that the order will matter when the workflow is applied!

```{r add-settings}
# Conversion of list to Workflow object with validation
workflow <- Workflow(workflow)

# Adds the workflow to the engine. The order matters!
ms$Workflow <- workflow

# Printing the data processing workflow
show(ms$Workflow)
```

# `run_workflow()`

The *Workflow* can be applied by `run_workflow()`, as demonstrated below. Note that with `run_workflow()`, the processing modules are applied with the same order as they were added.

```{r do-workflow, results='hide'}
# Runs all ProcessingStep added
ms$run_workflow()
```

```{r}

get_patRoon_MSPeakLists(ms$Results$MassSpecResults_NonTargetAnalysis)


get_suspects(ms$Results$MassSpecResults_NonTargetAnalysis)

nts <- ms$Results$MassSpecResults_NonTargetAnalysis
fts <- get_features(nts, mass = db, ppm = 10, sec = 15)
fts
fts$quality[[17]]
```

# Results

The created features and feature groups can be inspected as `data.table` objects or plotted by dedicated methods of the [MassSpecResults_NonTargetAnalysis](https://odea-project.github.io/StreamFind/reference/MassSpecResults_NonTargetAnalysis.html) class. Internally, the *MassSpecEngine* stores the results in the results list of the [MassSpecAnalyses](https://odea-project.github.io/StreamFind/reference/MassSpecAnalyses.html) object, which can be accessed with `MassSpecEngine$Results$MassSpecResults_NonTargetAnalysis` as shown below. The *MassSpecResults_NonTargetAnalysis* object can be used for advanced operations, such as exporting the results to a database or other formats or use the native objects from other packages (e.g., [patRoon](https://github.com/rickhelmus/patRoon)) as we demonstrate further in this article.

```{r results-access}
# Accessing the MassSpecResults_NonTargetAnalysis from the Engine$Results
nts <- ms$Results$MassSpecResults_NonTargetAnalysis

class(nts)
```

## `data.table` objects

The features and feature groups can be obtained as `data.table` with the `get_features()` and `get_groups()` methods. The methods also allow to look for specific features/feature groups using mass, mass-to-charge ratio, retention time and drift time targets, as show below for a small set of compound targets where mass and retention time expected values are known. Note that drift time is only applicable for MS data with ion mobility separation.

```{r database-show}
db
```

```{r show-target-groups}
# Compounds are searched by monoisotopic mass and retention time
# ppm and sec set the mass (im ppm) and time (in seconds) allowed deviation, respectively
# average applies a mean to the intensities in each analysis replicate group
get_groups(nts, mass = db, ppm = 10, sec = 15, average = TRUE)
```

Already by inspection of the `data.table`, it is possible to see compounds detected in the influent but not in the effluent (e.g., Carbamazepine) or compounds that are appear to be reduced during ozonation (e.g., Metoprolol). Since positive and negative ionization mode were combined, there are compounds that appear in both polarities and are grouped by neutral monoisotopic mass (e.g., Candesartan and Diclofenac).

## `plot_groups` methods

For a better overview of the results, the method `plot_groups()` or even more detailed the method `plot_groups_overview()` can be used.

```{r plot-groups}
# set legendNames to TRUE for using the names in db as legend
plot_groups(nts, mass = db, ppm = 10, sec = 15, legendNames = TRUE)
```

```{r plot-groups-overview, fig.height=10}
plot_groups_overview(nts, mass = db, ppm = 5, sec = 10, legendNames = TRUE)
```

## Filtered not removed

The *FilterFeatures* method was applied to filter features according to defined conditions/thresholds. Yet, the filtered features were not removed but just tagged as filtered. For instance, when the method `MassSpecEngine$get_groups()` is run with `filtered` argument set to `TRUE`, the filtered features are also shown. Below, we search for the same compounds as above but with the `filtered` argument set to `TRUE`. Potential features from Valsartan are now returned but were filtered due to low intensity. Note that when extracting features based on basic parameters, i.e. mass and time, does not mean that features are identified. The identification of features is a more complex process and requires additional information, such as MS/MS data as in the processing method suspect screening.

```{r show-filtered-is}
# Set filtered to TRUE for showing filtered features/feature groups
get_groups(nts, mass = db, ppm = 5, sec = 10, average = TRUE, filtered = TRUE)
```

## Internal Standards

The method *FindInternalStandards* was applied for tagging spiked internal standards and the results can be obtained with the dedicated method `MassSpecEngine$get_internal_standards()` or plotted as a quality overview using the method `MassSpecEngine$plot_internal_standards()`, as shown below. The plot gives an overview of the mass, retention time and intensity variance of the internal standards across the analyses in the project.

```{r database-is-show}
# List of spiked internal standards
db_is
```

```{r get_is}
# Gets the internal standards evaluation data.table
get_internal_standards(nts)[, 1:7]
```

```{r plot_is_qc, fig.height=10, fig.cap="Quality control of spiked internal standards"}
plot_internal_standards(nts)
```

```{r plot_is_profile, fig.cap="Internal standards profile across analyses"}
plot_groups_profile(nts, mass = db_is, ppm = 8, sec = 10, filtered = TRUE, legendNames = TRUE)
```

## Components

The method *AnnotateFeatures* was applied to annotate the natural isotopes and adducts into components. Implementation of annotation for in-source fragments is planned but not yet available with the StreamFind algorithm. The method `MassSpecEngine$get_components()` can be used to search for components, as shown below for the analysis number 11. Because the filters `excludeIsotopes` and `minIntensity` were applied, the isotopic features are likely filtered.

```{r show-components}
# Components of Diclofenac and Candesartan in analysis 11
components_example <- get_components(
  nts,
  analyses = 11,
  mass = db[db$name %in% c("Diclofenac", "Candesartan"), ],
  ppm = 5, sec = 10,
  filtered = TRUE
)

# Subset of the components data.table
components_example[, c(2:6, 8, 30:34)]
```

The components (i.e., isotopes and adducts) can also be visualized with the method `MassSpecEngine$map_components()`, as shown below for the internal standards added to analysis 11. Note that again the `filtered` argument was set to `TRUE` to return also filtered features, as the internal standards were likely excluded by blank subtraction.

```{r plot-components, fig.cap="Components (i.e., isotopes and adducts) of internal standards in analysis 11"}
map_components(
  nts,
  analyses = 11,
  mass = db_is,
  ppm = 8, sec = 10,
  filtered = TRUE,
  legendNames = TRUE
)
```

## Suspects

The methods `MassSpecEngine$get_suspects()` and `MassSpecEngine$plot_suspects()` can be used to inspect the suspect screening results. In the plot function, a second plot is added to compare the experimental fragmentation pattern (top) with the fragmentation pattern of the respective reference standard (down) added within the database. The `colorBy` argument can be set to *targets+replicates* to legend the plot with combined keys of suspect target names and analysis replicate names.

```{r get_suspects}
get_suspects(nts)[, c(1, 5, 14)]
```

```{r plot-suspects, fig.height=10}
plot_suspects(nts, colorBy = "targets+replicates")
```

## Methods from patRoon

The MassSpecResults_NonTargetAnalysis object holds methods to obtain original objects from the [patRoon](https://github.com/rickhelmus/patRoon) package. For instance, the S4 class `features` or `featureGroups` objects can be obtained via the `get_patRoon_features` method of the *MassSpecResults_NonTargetAnalysis* results. The `patRoon` package provides a comprehensive set of functions, as shown below. See more information in the [patRoon reference documentation](https://rickhelmus.github.io/patRoon/reference/index.html).

```{r pat-features}
# Native patRoon object
fGroups <- get_patRoon_features(nts, filtered = FALSE, featureGroups = TRUE)

fGroups
```

```{r pat-plot, fig.alt="UpSet plot of features"}
# Using the native patRoon's plotUpSet method
patRoon::plotUpSet(fGroups)
```

## Fold-change analysis

The method `get_fold_change()` and correspondent `plot_fold_change()` can be used to calculate and plot the fold-change between influent and effluent samples. The method calculates the fold-change for each feature group and replicates, as shown below. The plot shows the fold-change for each feature group across the replicates. The fold-change is calculated according to [Bader et al. (2017)](https://pubs.acs.org/doi/10.1021/acs.analchem.7b03037), leveraging the replicates variance to increase the significance of the fold-change. Formation is not in the plot as no new features were detected in the effluent of the wastewater ozonation treatment step.

```{r plot-fold-change}
plot_fold_change(
  nts,
  replicatesIn = c("influent_neg", "influent_pos"),
  replicatesOut = c("effluent_neg", "effluent_pos"),
  filtered = FALSE,
  constantThreshold = 0.5,
  eliminationThreshold = 0.2,
  correctIntensity = TRUE,
  fillZerosWithLowerLimit = FALSE, # set to TRUE for filling zeros with lower limit argument
  lowerLimit = 200,
  normalized = FALSE
)
```

Alternatively, the `plotVolcano` from [patRoon](https://github.com/rickhelmus/patRoon) can be used to plot fold-change. For more information check the [patRoon reference documentation](https://rickhelmus.github.io/patRoon/reference/index.html). Note that the function `patRoon::getFCParams` is used to define the parameters for the fold-change calculation. Below we use the default argument values, only the in and out replicate names are set.

```{r pat-volcano, fig.alt="Volcano plot of fold-change"}
patRoon::plotVolcano(
  fGroups, # obtained above with get_patRoon_features
  patRoon::getFCParams(c("influent_pos", "effluent_pos"))
)
```

# Compounds

```{r subset-pos}
# clone the MassSpecEngine to avoid overwriting
# R6 classes are reference classes so we need to clone to effectively copy
ms_pos <- ms$clone()

# subset with only positive analyses
ms_pos$Analyses <- ms_pos$Analyses[info(ms_pos$Analyses)$polarity == "1"]

# subsetting the Analyses, also subsets the Results
show(ms_pos$Results$MassSpecResults_NonTargetAnalysis)

# feature groups of interest for compounds generation showncase
grs <- get_groups(nts, mass = db, ppm = 10, sec = 15)[["group"]]
grs

# subsetting on feature groups
ms_pos$Results <- ms_pos$Results$MassSpecResults_NonTargetAnalysis[, grs]
show(ms_pos$Results$MassSpecResults_NonTargetAnalysis)
```

```{r get-compounds, results='asis'}
# Applies the generates compounds using MetFrag via the patRoon package
ms_pos$run(
  MassSpecMethod_GenerateCompounds_metfrag(
    method = "CL",
    database = "comptox"
  )
)
```

```{r}
# extract annotated compounds from the MassSpecResults_NonTargetAnalysis
get_compounds(ms_pos$Results$MassSpecResults_NonTargetAnalysis)[, c(1, 2, 3, 6, 7, 16, 20)]
```

# More to come

Future assets/functionalities:

- Annotation of in-source fragments
- Screening of transformation products using the *biotransformer* via patRoon

***
