---
title: 'Detection of Quality Deviations in Paracetamol Products'
subtitle: |
  | Fusion of Mass Spectrometry and Raman Spectroscopy Data
  | <br>
author:
  name: Ricardo Cunha
  email: cunha@iuta.de
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    fig_caption: true
    toc: true
    number_sections: true
    toc_float: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Detection of Quality Deviations in Paracetamol Products}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(plotly)
library(data.table)
library(StreamFind)

knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center", fig.width = 9, results = "markup", comment = ""
)

#main_dir <- "E:/iSoft/Paracetamol" # StreamFind Workstation

main_dir <- "C:/Users/apoli/Documents/iSoft/Paracetamol" # Ricardo laptop

ms_dir <- paste0(main_dir, "/ms/mzml")

ms_dir_extra <- paste0(main_dir, "/ms/extra/mzml")

ms_files <- list.files(ms_dir, pattern = ".mzML", full.names = TRUE)

ms_files_extra <- list.files(ms_dir_extra, pattern = ".mzML", full.names = TRUE)

raman_dir <- paste0(main_dir, "/raman")

raman_dir_extra <- paste0(main_dir, "/raman/extra")

raman_files <- list.files(raman_dir, pattern = ".asc", full.names = TRUE)

raman_files_extra <- list.files(raman_dir_extra, pattern = ".asc", full.names = TRUE)
```

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

<br>

<br>

***

This article demonstrates how StreamFind can be used to assess quality deviations of Fizamol paracetamol products using 
Mass Spectrometry (MS) combined with Raman Spectroscopy data. Each product was measured with MS and Raman. The data is 
pre-processed and fused to be used in unsupervised and supervised statistical analysis. Finally, the quality of the 
products is evaluated based on the statistical analysis. The quantity of paracetamol in the products is also estimated 
via statistic analysis.

# Pre-processing

The data is pre-processed using StreamFind's built-in data dedicated engines for MS and Raman data. Pre-processing 
workflows for each data type are executed before data fusion to improve statistic analysis.

## MS

```{r ms-setup, include=FALSE}
ms <- MassSpecEngine$new(analyses = c(ms_files, ms_files_extra[c(5:40, 49:60)]), headers = list(name = "BSA Quality Evaluation Project"))

ms$analyses$replicates <- c(
  rep("fiz_a1", 3),
  rep("fiz_a2", 3),
  rep("fiz_a3", 3),
  rep("fiz_b1", 3),
  rep("fiz_b2", 3),
  rep("fiz_b3", 3),
  rep("fiz_c1", 3),
  rep("fiz_c2", 3),
  rep("fiz_c3", 3),
  rep("fiz_caff1", 3),
  rep("fiz_caff2", 3),
  rep("fiz_caff3", 3),
  rep("fiz_ass1", 3),
  rep("fiz_ass2", 3),
  rep("fiz_ass3", 3),
  rep("fiz_d1", 3),
  rep("fiz_d2", 3),
  rep("fiz_d3", 3),
  rep("fiz_acid1", 3),
  rep("fiz_acid2", 3),
  rep("fiz_acid3", 3),
  rep("ass_n_1", 3),
  rep("ass_n_2", 3),
  rep("ass_n_3", 3),
  rep("ass_1", 3),
  rep("ass_2", 3),
  rep("ass_3", 3),
  rep("caff_1", 3),
  rep("caff_2", 3),
  rep("caff_3", 3),
  rep("par_1", 3),
  rep("par_2", 3),
  rep("par_3", 3),
  rep("blank", 23),
  rep("fiz_e01", 3),
  rep("fiz_e02", 3),
  rep("fiz_e03", 3),
  rep("fiz_e04", 3),
  rep("fiz_e05", 3),
  rep("fiz_e06", 3),
  rep("fiz_e07", 3),
  rep("fiz_e08", 3),
  rep("fiz_e09", 3),
  rep("fiz_e10", 3),
  rep("fiz_e11", 3),
  rep("fiz_e12", 3),
  rep("blank_e", 12)
)

ms$analyses <- ms$analyses[c(1:63, 73:99, 118:121, 123:158, 161:170)]

ms$analyses$blanks <- c(rep("blank", 94), rep("blank_e", 140 - 94))

ms$load_spectra()
```

```{r ms-overview, echo=FALSE}
ov <- ms$analyses$info[, 1:3]
data.table::setorder(ov, replicate)
ov$description <- c(
  rep("Aspirin standard", 9),
  rep("Reference background", 4),
  rep("Reference background", 10),
  rep("Caffeine standard", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol with acetic acid", 9),
  rep("Paracetamol with aspirin", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol with caffeine", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol product", 36),
  rep("Paracetamol standard", 9)
)

kable(ov, caption = "Overview of MS analyses and respective analysis replicate groups and blanks.") %>%
kable_styling(font_size = 12, bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = TRUE) %>%
scroll_box(width = "100%", height = "400px")
```

<br>

```{r ms-workflow}
ms_workflow <- StreamFind::Workflow(
  list(
    MassSpecSettings_BinSpectra_StreamFind(binNames = c("rt", "mz"), binValues = c(5, 2), refBinAnalysis = 1),
    MassSpecSettings_NormalizeSpectra_minmax(),
    MassSpecSettings_AverageSpectra_StreamFind(),
    MassSpecSettings_SubtractBlankSpectra_StreamFind(negativeToZero = TRUE),
    MassSpecSettings_NormalizeSpectra_blockweight()
  )
)

ms$workflow <- ms_workflow

ms$print_workflow()
```

```{r ms-run-workflow}
ms$run_workflow()
```

```{r plot-ms-spectra, fig.cap="Pre-processed spectra for each analysis. Retention time dimension in he created bins is collapsed for each mass-to-charge ratio"}
ms$plot_spectra()
```

## Raman

```{r raman-setup, include=FALSE}
raman <- RamanEngine$new(analyses = c(raman_files, raman_files_extra))

raman$analyses$replicates <- c(
  rep("ass_1", 3),
  rep("ass_2", 3),
  rep("ass_3", 3),
  rep("caff_1", 3),
  rep("caff_2", 3),
  rep("caff_3", 3),
  rep("blank", 9),
  rep("fiz_c1", 3),
  rep("fiz_c2", 3),
  rep("fiz_c3", 3),
  rep("fiz_ass_n_1", 3),
  rep("fiz_ass_n_2", 3),
  rep("fiz_ass_n_3", 3),
  rep("fiz_ass1", 3),
  rep("fiz_ass2", 3),
  rep("fiz_ass3", 3),
  rep("fiz_caff1", 3),
  rep("fiz_caff2", 3),
  rep("fiz_caff3", 3),
  rep("fiz_a1", 3),
  rep("fiz_a2", 3),
  rep("fiz_a3", 3),
  rep("fiz_d1", 3),
  rep("fiz_d2", 3),
  rep("fiz_d3", 3),
  rep("fiz_b1", 3),
  rep("fiz_b2", 3),
  rep("fiz_b3", 3),
  rep("background", 4),
  rep("blank_e", 9),
  rep("fiz_acid1", 3),
  rep("fiz_acid2", 3),
  rep("fiz_acid3", 3),
  rep("fiz_e01", 3),
  rep("fiz_e02", 3),
  rep("fiz_e03", 3),
  rep("fiz_e04", 3),
  rep("fiz_e05", 3),
  rep("fiz_e06", 3),
  rep("fiz_e07", 3),
  rep("fiz_e08", 3),
  rep("fiz_e09", 3),
  rep("fiz_e10", 3),
  rep("fiz_e11", 3),
  rep("fiz_e12", 3),
  rep("par_1", 3),
  rep("par_odd", 3),
  rep("par_3", 3),
  rep("par_2", 3),
  rep("cali_150_mgml", 9),
  rep("cali_010_mgml", 9),
  rep("cali_025_mgml", 9),
  rep("cali_050_mgml", 9),
  rep("cali_075_mgml", 9),
  rep("background", 4)
)

raman$analyses <- raman$analyses[c(1:36, 46:90, 95:151, 155:160)]
raman$analyses$blanks <- c(rep("blank", 81), rep("blank_e", 9), rep("blank", 9), rep("blank_e", 36), rep("blank", 9))
```

```{r raman-overview, echo=FALSE}
raman_ov <- raman$get_overview()[, 1:3]
data.table::setorder(raman_ov, replicate)
raman_ov$description <- c(
  rep("Aspirin standard", 9),
  rep("Reference background", 9),
  rep("Reference background", 9),
  rep("Caffeine standard", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol with acetic acid", 9),
  rep("Paracetamol with aspirin", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol with caffeine", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol product", 36),
  rep("Paracetamol standard", 9)
)

kable(raman_ov, caption = "Overview of MS analyses and respective analysis replicate groups and blanks.") %>%
kable_styling(font_size = 12, bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = TRUE) %>%
scroll_box(width = "100%", height = "400px")
```

```{r raman-workflow}
raman_workflow <- StreamFind::Workflow(
  list(
    RamanSettings_AverageSpectra_StreamFind(),
    RamanSettings_SubtractBlankSpectra_StreamFind(),
    RamanSettings_SmoothSpectra_savgol(fl = 21, forder = 4, dorder = 0),
    RamanSettings_DeleteSpectraSection_StreamFind(shiftmin = -100, shiftmax = 350),
    RamanSettings_DeleteSpectraSection_StreamFind(shiftmin = 1800, shiftmax = 2500),
    RamanSettings_CorrectSpectraBaseline_airpls(lambda = 45, differences = 1, itermax = 30),
    RamanSettings_NormalizeSpectra_minmax(),
    RamanSettings_NormalizeSpectra_blockweight()
  )
)

raman$workflow <- raman_workflow

raman$print_workflow()
```

```{r raman-run-workflow}
raman$run_workflow()
```

```{r plot-raman-spectra, fig.cap="Pre-processed Raman spectra for analysis replicate."}
raman$plot_spectra()
```

# Fusion

The statistical analysis is performed on the fused data for combining orthogonal information in MS and Raman data.

```{r fusion}
ms_spec_mat <- ms$get_spectra_matrix()
raman_spec_mat <- raman$get_spectra_matrix()
fused_mat <- cbind(ms_spec_mat, raman_spec_mat)
# The fused matrix is mean centered using the mdatools R package
fused_mat <- mdatools::prep.autoscale(fused_mat, center = TRUE, scale = FALSE)
str(fused_mat)
```

# Statistic analysis

## Unsupervised

Principle Components Analysis (PCA) can be used for unsupervised evaluation, meaning that we assume that the main 
(or train) set of analyses are of good quality and we test new analyses to evaluate the proximity to the main set. 
One could say that it is a supervised approach but in the sense that we assume the quality of the main set. Yet, we are 
not classifying the possible deviation of the new analyses which would be a clear supervised approach, as we show in 
the next sub chapter.

```{r sets-show}
main_set_idx <- c(1:3, 4:5)

main_set <- fused_mat[main_set_idx, ]
rownames(main_set)

test_set <- fused_mat[-main_set_idx, ]
rownames(test_set)
```

```{r pca-engine, fig.cap="PCA model (native plot from mdatools R package).", fig.height=9, warning=FALSE}
stat <- StatisticEngine$new(analyses = main_set)
# The alpha and gamma parameters are used to evaluate the model limits
stat$run(StatisticSettings_MakeModel_pca_mdatools(ncomp = 3, alpha = 0.01, gamma = 0.001))
plot(stat$model, showText = TRUE)
```

```{r pca-predict, warning=FALSE, fig.cap="PCA model prediction distances. For more information about the distances plese consult the mdatools guide in https://mdatools.com/docs/pca--distances-and-limits.html."}
stat$predict(test_set)
plot(stat$model, showText = TRUE)
```

## Supervised (classification)

The k-nearest neighbors (KNN) can be used for classification of analyses based on pre-classified training analyses.

```{r sets-show-knn}
train_set <- fused_mat[unique(c(grep("1", rownames(fused_mat)), grep("2", rownames(fused_mat)))), ]
train_set <- train_set[order(rownames(train_set)), ]
rownames(train_set)

test_set <- fused_mat[unique(grep("3|4|5|6|7|8|9", rownames(fused_mat))), ]
rownames(test_set)
```

```{r knn-classes}
labels <- c(rep("ass", 2), rep("caff", 2), rep("fiz", 2), rep("fiz_acid", 2), rep("fiz_ass", 2),
  rep("fiz", 2), rep("fiz", 2), rep("fiz_caff", 2), rep("fiz", 7), rep("par", 2))
```

```{r knn-engine}
stat2 <- StatisticEngine$new(analyses = train_set)
stat2$add_classes(labels)
stat2$run(StatisticSettings_MakeModel_knn(k = 3))
stat2$predict(test_set)
stat2$model$prediction$results
```
