---
title: 'Identification of Intact Monoclonal Antibody with LC-HRMS'
author:
  name: Ricardo Cunha
  email: cunha@iuta.de
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    fig_caption: true
    toc: true
    number_sections: true
    toc_float: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{'Identification of Intact Monoclonal Antibody with LC-HRMS'}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(plotly)
library(data.table)
library(StreamFind)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 9,
  results = "markup",
  comment = ""
)

options(
  DT.options = list(
    dom = "lfrtip",
    autoWidth = TRUE,
    pageLength = 10,
    lengthMenu = c(10, 20, 50, 75, 100),
    initComplete = htmlwidgets::JS(
      "function(settings, json) {",
      "$(this.api().table().body()).css({'font-size': '12px'});",
      "$(this.api().table().body()).css({'white-space': 'nowrap'});",
      "}"
    )
  )
)

path <- "C:/Users/apoli/Documents/example_files/isoft/240819_BVCZ"
files <- list.files(path = path, pattern = "\\.d$", full.names = TRUE)[c(6:8, 16:18)]
```

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

<br>

<br>

***

# Introduction

This document demonstrates the identification of the intact monoclonal antibody Bevacizumab using LC-HRMS data. The files consist of a triplicate QC and a triplicate sample to be identified.

```{r files}
basename(files)
```

# Processing Engine

The `StreamFind` package provides a flexible and powerful framework for processing mass spectrometry data. In this example, we will use the `MassSpecEngine` class to handle the LC-HRMS data. The engine will be initialized with the raw files, and we will specify that the data should be centroided and processed at level 1. Note that The `msconvert` from [ProteoWizard](https://proteowizard.sourceforge.io/) was used in the background to convert the files to the open format mzML.

```{r engine}
# Starts engine for identification
ms <- StreamFind::MassSpecEngine$new(
  metadata = list(name = "Identification of Bevacizumab"),
  analyses = files,
  centroid = TRUE,
  levels = 1
)

# Assign analysis replicate groups
ms$Analyses$replicates <- c(
  rep("Sample", 3),
  rep("QC_2.5", 3)
)

ms$Analyses$info[, c(1:2, 4, 5)]
```

# Workflow

```{r load-tic}
# Load Total Ion Chromatograms (TIC) for detecting the API
ms$run(MassSpecMethod_LoadChromatograms_native(chromatograms = "TIC"))

# Smooth TIC chromatograms
ms$run(MassSpecMethod_SmoothChromatograms_movingaverage(windowSize = 5))
```

```{r plot-tic}
ms$plot_chromatograms(colorBy = "replicates")
```

```{r find-tic-peaks}
# Find peaks in the TIC chromatograms
ms$run(
  MassSpecMethod_FindChromPeaks_LocalMaxima(
    minWidth = 5,
    maxWidth = 10,
    minHeight = 10E6
  )
)
```

```{r plot-tic-peaks}
ms$plot_chromatograms_peaks(colorBy = "replicates+targets")
```

```{r load-spectra}
# Load spectra based on integrated chromatograms
ms$run(
  MassSpecMethod_LoadSpectra_chrompeaks(
    mzmin = 2500,
    mzmax = 3700,
    levels = 1,
    minIntensity = 1000
  )
)
```

```{r plot-spectra, fig.alt="plot-spectra-static"}
# plots only the spectra of the first analysis
plot_spectra(ms$Spectra, analyses = 1, interactive = FALSE)
```

```{r spectra-charges}
# Calculate spectra charges for de-clustering and deconvolution
ms$run(
  MassSpecMethod_CalculateSpectraCharges_native(
    onlyTopScans = TRUE,
    topScans = 6,
    roundVal = 25,
    relLowCut = 0.2,
    absLowCut = 0,
    top = 10
  )
)
```

```{r plot-spectra-charges}
plot_spectra_charges(ms$Spectra)
```

```{r deconvolute-mass}
# Deconvolute spectra to obtain mass values
ms$run(
  MassSpecMethod_DeconvoluteSpectra_native(
    clustVal = 0.001,
    window = 25
  )
)

# Smooth deconvoluted spectra
ms$run(
  MassSpecMethod_SmoothSpectra_savgol(
    fl = 141,
    forder = 4,
    dorder = 0
  )
)
```

```{r plot-spectra-deconvoluted, fig.alt="plot-spectra-static-deconvoluted"}
plot_spectra(ms$Spectra, analyses = 1, interactive = FALSE)
```

```{r spectra-maxima}
# Find spectra peaks to verify mass values and evaluate identity
ms$run(
  MassSpecMethod_FindSpectraMaxima_native(
    minWidth = 10,
    maxWidth = 60,
    minHeight = 50000
  )
)
```

```{r plot-spectra-maxima, fig.alt="plot-spectra-static-maxima"}
# Plot spectra peaks for the first analysis
plot_spectra_peaks(
  ms$Spectra,
  analyses = 1,
  colorBy = "analyses+targets",
  interactive = FALSE
)
```
