---
title: 'Detection of quality deviations in paracetamol products'
subtitle: |
  | Fusion of Mass Spectrometry and Raman Spectroscopy Data
  | <br>
author:
  name: Ricardo Cunha
  email: cunha@iuta.de
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    fig_caption: true
    toc: true
    number_sections: true
    toc_float: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Fusion of Mass Spectrometry and Raman Spectroscopy Data}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(plotly)
library(data.table)
library(StreamFind)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center", fig.width = 9, results = "markup", comment = "")

#main_dir <- "E:/iSoft/Paracetamol" # StreamFind Workstation

main_dir <- "C:/Users/apoli/Documents/iSoft/Paracetamol" # Ricardo laptop

ms_dir <- paste0(main_dir, "/ms/mzml")

ms_dir_extra <- paste0(main_dir, "/ms/extra/mzml")

ms_files <- list.files(ms_dir, pattern = ".mzML", full.names = TRUE)

ms_files_extra <- list.files(ms_dir_extra, pattern = ".mzML", full.names = TRUE)

raman_dir <- paste0(main_dir, "/raman")

raman_dir_extra <- paste0(main_dir, "/raman/extra")

raman_files <- list.files(raman_dir, pattern = ".asc", full.names = TRUE)

raman_files_extra <- list.files(raman_dir_extra, pattern = ".asc", full.names = TRUE)
```

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

<br>

<br>

***

This article demonstrates how StreamFind can be used to assess quality deviations of Fizamol paracetamol products using Mass Spectrometry (MS) combined with Raman Spectroscopy data. Each product was measured with MS and Raman. The data is pre-processed and fused to be used in unsupervised and supervised statistical analysis. Finally, the quality of the products is evaluated based on the statistical analysis. The quantity of paracetamol in the products is also estimated via statistic analysis.

# Pre-processing

The data is pre-processed using StreamFind's built-in data dedicated engines for MS and Raman data. Pre-processing workflows for each data type are executed before data fusion to improve statistic analysis.

## MS

```{r ms-setup, include=FALSE}
ms <- MassSpecEngine$new(files = c(ms_files, ms_files_extra[c(5:40, 49:60)]), headers = list(name = "BSA Quality Evaluation Project"))

ms$add_replicate_names(
  c(rep("fiz_a1", 3),
    rep("fiz_a2", 3),
    rep("fiz_a3", 3),
    rep("fiz_b1", 3),
    rep("fiz_b2", 3),
    rep("fiz_b3", 3),
    rep("fiz_c1", 3),
    rep("fiz_c2", 3),
    rep("fiz_c3", 3),
    rep("fiz_caff1", 3),
    rep("fiz_caff2", 3),
    rep("fiz_caff3", 3),
    rep("fiz_ass1", 3),
    rep("fiz_ass2", 3),
    rep("fiz_ass3", 3),
    rep("fiz_d1", 3),
    rep("fiz_d2", 3),
    rep("fiz_d3", 3),
    rep("fiz_acid1", 3),
    rep("fiz_acid2", 3),
    rep("fiz_acid3", 3),
    rep("ass_n_1", 3),
    rep("ass_n_2", 3),
    rep("ass_n_3", 3),
    rep("ass_1", 3),
    rep("ass_2", 3),
    rep("ass_3", 3),
    rep("caff_1", 3),
    rep("caff_2", 3),
    rep("caff_3", 3),
    rep("par_1", 3),
    rep("par_2", 3),
    rep("par_3", 3),
    rep("blank", 23),
    rep("fiz_e01", 3),
    rep("fiz_e02", 3),
    rep("fiz_e03", 3),
    rep("fiz_e04", 3),
    rep("fiz_e05", 3),
    rep("fiz_e06", 3),
    rep("fiz_e07", 3),
    rep("fiz_e08", 3),
    rep("fiz_e09", 3),
    rep("fiz_e10", 3),
    rep("fiz_e11", 3),
    rep("fiz_e12", 3),
    rep("blank_e", 12)
  )
)

ms <- ms$subset_analyses(analyses = c(1:63, 73:99, 118:121, 123:158, 161:170))

ms$add_blank_names(c(rep("blank", 94), rep("blank_e", 140 - 94)))

ms$load_spectra()
```

```{r ms-overview, echo=FALSE}
ov <- ms$get_overview()[, 1:3]
data.table::setorder(ov, replicate)
ov$description <- c(
  rep("Aspirin standard", 9),
  rep("Reference background", 4),
  rep("Reference background", 10),
  rep("Caffeine standard", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol with acetic acid", 9),
  rep("Paracetamol with aspirin", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol with caffeine", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol product", 36),
  rep("Paracetamol standard", 9)
)

kable(ov, caption = "Overview of MS analyses and respective analysis replicate groups and blanks.") %>%
kable_styling(font_size = 12, bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = TRUE) %>%
scroll_box(width = "100%", height = "400px")
```

<br>

```{r ms-workflow}
ms_workflow_settings <- list(
  MassSpecSettings_BinSpectra_StreamFind(bins = list("rt" = 5, "mz" = 2), refBinAnalysis = 1),
  MassSpecSettings_NormalizeSpectra_minmax(),
  MassSpecSettings_AverageSpectra_StreamFind(),
  MassSpecSettings_SubtractBlankSpectra_StreamFind(negativeToZero = TRUE),
  MassSpecSettings_NormalizeSpectra_blockweight()
)

ms$add_settings(ms_workflow_settings)

ms$print_workflow()
```

```{r ms-run-workflow}
ms$run_workflow()
```

```{r plot-ms-spectra, fig.cap="Pre-processed spectra for each analysis. Retention time dimension in he created bins is collapsed for each mass-to-charge ratio"}
ms$plot_spectra()
```

```{r ms-spectra}
ms$get_spectra()[1:10, ]
```

## Raman

```{r raman-setup, include=FALSE}
raman_all <- RamanEngine$new(c(raman_files, raman_files_extra))

raman_all$add_replicate_names(
  c(
    rep("ass_1", 3),
    rep("ass_2", 3),
    rep("ass_3", 3),
    rep("caff_1", 3),
    rep("caff_2", 3),
    rep("caff_3", 3),
    rep("blank", 9),
    rep("fiz_c1", 3),
    rep("fiz_c2", 3),
    rep("fiz_c3", 3),
    rep("fiz_ass_n_1", 3),
    rep("fiz_ass_n_2", 3),
    rep("fiz_ass_n_3", 3),
    rep("fiz_ass1", 3),
    rep("fiz_ass2", 3),
    rep("fiz_ass3", 3),
    rep("fiz_caff1", 3),
    rep("fiz_caff2", 3),
    rep("fiz_caff3", 3),
    rep("fiz_a1", 3),
    rep("fiz_a2", 3),
    rep("fiz_a3", 3),
    rep("fiz_d1", 3),
    rep("fiz_d2", 3),
    rep("fiz_d3", 3),
    rep("fiz_b1", 3),
    rep("fiz_b2", 3),
    rep("fiz_b3", 3),
    rep("background", 4),
    rep("blank_e", 9),
    rep("fiz_acid1", 3),
    rep("fiz_acid2", 3),
    rep("fiz_acid3", 3),
    rep("fiz_e01", 3),
    rep("fiz_e02", 3),
    rep("fiz_e03", 3),
    rep("fiz_e04", 3),
    rep("fiz_e05", 3),
    rep("fiz_e06", 3),
    rep("fiz_e07", 3),
    rep("fiz_e08", 3),
    rep("fiz_e09", 3),
    rep("fiz_e10", 3),
    rep("fiz_e11", 3),
    rep("fiz_e12", 3),
    rep("par_1", 3),
    rep("par_odd", 3),
    rep("par_3", 3),
    rep("par_2", 3),
    rep("cali_150_mgml", 9),
    rep("cali_010_mgml", 9),
    rep("cali_025_mgml", 9),
    rep("cali_050_mgml", 9),
    rep("cali_075_mgml", 9),
    rep("background", 4)
  )
)

raman <- raman_all$subset_analyses(analyses = c(1:36, 46:90, 95:151, 155:160))
raman$add_blank_names(c(rep("blank", 81), rep("blank_e", 9), rep("blank", 9), rep("blank_e", 36), rep("blank", 9)))
```

```{r raman-overview, echo=FALSE}
raman_ov <- raman$get_overview()[, 1:3]
data.table::setorder(raman_ov, replicate)
raman_ov$description <- c(
  rep("Aspirin standard", 9),
  rep("Reference background", 9),
  rep("Reference background", 9),
  rep("Caffeine standard", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol with acetic acid", 9),
  rep("Paracetamol with aspirin", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol with caffeine", 9),
  rep("Paracetamol product", 9),
  rep("Paracetamol product", 36),
  rep("Paracetamol standard", 9)
)

kable(raman_ov, caption = "Overview of MS analyses and respective analysis replicate groups and blanks.") %>%
kable_styling(font_size = 12, bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = TRUE) %>%
scroll_box(width = "100%", height = "400px")
```

```{r raman-workflow}
raman_workflow_settings <- list(
  RamanSettings_AverageSpectra_StreamFind(),
  RamanSettings_SubtractBlankSpectra_StreamFind(),
  RamanSettings_SmoothSpectra_savgol(fl = 21, forder = 4, dorder = 0),
  RamanSettings_DeleteSpectraSection_StreamFind(list("shift" = c(-100, 350))),
  RamanSettings_DeleteSpectraSection_StreamFind(list("shift" = c(1800, 2500))),
  RamanSettings_CorrectSpectraBaseline_airpls(lambda = 45, differences = 1, itermax = 30),
  RamanSettings_NormalizeSpectra_minmax(),
  RamanSettings_NormalizeSpectra_blockweight()
)

raman$add_settings(raman_workflow_settings)

raman$print_workflow()
```

```{r raman-run-workflow}
raman$run_workflow()
```

```{r plot-raman-spectra, fig.cap="Pre-processed Raman spectra for analysis replicate."}
raman$plot_spectra()
```

```{r raman-spectra}
raman$get_spectra()[1:10, ]
```

# Fusion

The statistical analysis is performed on the fused data for combining orthogonal information in MS and Raman data.

```{r fusion}
ms_spec_mat <- ms$get_spectra_matrix()
raman_spec_mat <- raman$get_spectra_matrix()
fused_mat <- cbind(ms_spec_mat, raman_spec_mat)
# the matrix is mean centered using the mdatools R package
fused_mat <- mdatools::prep.autoscale(fused_mat, center = TRUE, scale = FALSE)
```

```{r plot-fused, include=FALSE, fig.cap="Fused spectra."}
cl <- StreamFind:::.get_colors(rownames(fused_mat))
fig <- plot_ly()
xVal <- seq_len(length(fused_mat[1, ]))
for (i in seq_len(nrow(fused_mat))) {
  fig <- fig %>% add_trace(
    x = xVal,
    y = fused_mat[i, ],
    type = "scatter", mode = "lines",
    line = list(width = 0.5, color = unname(cl[i])),
    name = names(cl)[i],
    legendgroup = names(cl)[i],
    showlegend = TRUE
  )
}
xaxis <- list(
  linecolor = toRGB("black"),
  linewidth = 2, title = "Bins",
  titlefont = list(size = 12, color = "black")
)
yaxis <- list(
  linecolor = toRGB("black"),
  linewidth = 2, title = "Intensity",
  titlefont = list(size = 12, color = "black")
)
fig <- fig %>% plotly::layout(xaxis = xaxis, yaxis = yaxis)
fig
```

# Statistic analysis

## Unsupervised

Principle Components Analysis (PCA) can be used for unsupervised evaluation, meaning that we assume that the main (or train) set of analyses are of good quality and we test new analyses to evaluate the proximity to the main set. One could say that it is a supervised approach but in the sense that we assume the quality of the main set. Yet, we are not classifying the possible deviation of the new analyses which would be a clear supervised approach, as we show in the next sub chapter.

```{r sets-show}
main_set <- fused_mat[c(7:9, 16:21, 25:33), ]
rownames(main_set)

test_set <- fused_mat[c(1:6, 10:15, 22:24, 34:42), ]
rownames(test_set)
```

```{r pca-engine, fig.cap="PCA model (native plot from mdatools R package).", fig.height=9}
stat <- StatisticEngine$new(data = main_set)
# the alpha and gamma parameters are used to evaluate the model limits
stat$process(StatisticSettings_MakeModel_pca_mdatools(ncomp = 3, alpha = 1E-4, gamma = 1E-5))
# original plot from mdatools R package
plot(stat$model)
```

```{r pca-predict, fig.cap="PCA model prediction distances. For more information about the distances plese consult the mdatools guide in https://mdatools.com/docs/pca--distances-and-limits.html."}
stat$predict(test_set)
stat$plot_predicted_distances()
```

## Supervised (classification)

The k-nearest neighbors (KNN) can be used for classification of analyses based on pre-classified training analyses.

```{r sets-show-knn}
train_set <- fused_mat[unique(c(grep("1", rownames(fused_mat)), grep("2", rownames(fused_mat)))), ]
train_set <- train_set[order(rownames(train_set)), ]
rownames(train_set)

test_set <- fused_mat[unique(grep("3|4|5|6|7|8|9", rownames(fused_mat))), ]
rownames(test_set)
```

```{r knn-classes}
labels <- c(rep("ass", 2), rep("caff", 2), rep("fiz", 2), rep("fiz_acid", 2), rep("fiz_ass", 2),
  rep("fiz", 2), rep("fiz", 2), rep("fiz_caff", 2), rep("fiz", 7), rep("par", 2))
```

```{r knn-engine}
stat2 <- StatisticEngine$new(data = train_set)
stat2$add_classes(labels)
stat2$process(StatisticSettings_PrepareClassification_knn(k = 3))
stat2$classify(test_set)
stat2$classification_results
```

## Only MS

```{r pca-only-ms}
ms_spec_mat <- mdatools::prep.autoscale(ms_spec_mat, center = TRUE, scale = FALSE)
main_set_ms <- ms_spec_mat[c(7:9, 16:21, 25:33), ]
test_set_ms <- ms_spec_mat[c(1:6, 10:15, 22:24, 34:42), ]
stat_ms <- StatisticEngine$new(data = main_set_ms)
stat_ms$process(StatisticSettings_MakeModel_pca_mdatools(ncomp = 3, alpha = 1E-4, gamma = 1E-5))
```

```{r plot-pca-only-ms, fig.cap="PCA model for MS data only (native plot from mdatools R package).", fig.height=9}
plot(stat_ms$model)
```

```{r pca-omly-ms-predict, fig.cap="PCA model prediction distances for only MS data."}
stat_ms$predict(test_set_ms)
stat_ms$plot_predicted_distances()
```

Using MS data only for evaluation of quality deviations would not separate completely the analyses when compared to fused MS and Raman data. Thus, the contribution of Raman spectroscopy to explain structural deviations brings an advantage for routine quality evaluation of paracetamol and other pharmaceuticals. 

## Only Raman

```{r pca-only-raman}
raman_spec_mat <- mdatools::prep.autoscale(raman_spec_mat, center = TRUE, scale = FALSE)
main_set_raman <- raman_spec_mat[c(7:9, 16:21, 25:33), ]
test_set_raman <- raman_spec_mat[c(1:6, 10:15, 22:24, 34:42), ]
stat_raman <- StatisticEngine$new(data = main_set_raman)
stat_raman$process(StatisticSettings_MakeModel_pca_mdatools(ncomp = 3, alpha = 1E-4, gamma = 1E-5))
```

```{r plot-pca-only-raman, fig.cap="PCA model for Raman data only (native plot from mdatools R package).", fig.height=9}
plot(stat_raman$model)
```

```{r pca-omly-raman-predict, fig.cap="PCA model prediction distances for only Raman data."}
stat_raman$predict(test_set_raman)
stat_raman$plot_predicted_distances()
```

In the case of only using Raman data for evaluation of quality deviations, the separation of the analyses is clearer than when using fused MS and Raman data. This is due to the fact that the Raman data is more sensitive to structural deviations in the analyses. The lack of aspirin and acetic acid signals in the MS data probably contributed to the less clear separation of the analyses when using only MS data. On the contrary, because the signal of caffeine (*m/z* 194 Da) was visible in the MS data, a separation of the caffeine contaminated analyses was obtained also for MS only data. MS data is more precise when the goal is identification of the active product ingredient (i.e, paracetamol) therefore, the fusion of MS and Raman data is recommended for quality evaluation of pharmaceuticals as both techniques are complementary.

# Quantification with Raman

```{r ov-raman-conc, include=FALSE}
raman_quant <- raman_all$subset_analyses(analyses = c(19:27, 161:205, 155:157))
raman_quant$add_blank_names(rep("blank", 57))

ov_quant <- data.table::copy(raman_quant$get_overview()[, 1:3])
data.table::setorder(ov_quant, replicate)
ov_quant$concentration <- as.numeric(c(rep(0, 9), sub(".*?(\\d+).*", "\\1", ov_quant$replicate[10:nrow(ov_quant)]))) / 10
ov_quant$concentration[ov_quant$replicate %in% "par_3"] <- NA_real_

kable(ov_quant, caption = "Overview of Raman analyses for quantification with respective concentration.") %>%
kable_styling(font_size = 12, bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = TRUE) %>%
scroll_box(width = "100%", height = "400px")
```

```{r plot-raman-conc-raw, fig.cap="Raw averaged Raman spectra for each analysis replicate."}
raman_quant$plot_spectra(colorBy = "replicates")
```

```{r raman-conc-workflow}
raman_workflow_settings <- list(
  RamanSettings_AverageSpectra_StreamFind(),
  RamanSettings_SubtractBlankSpectra_StreamFind(),
  RamanSettings_SmoothSpectra_savgol(fl = 21, forder = 4, dorder = 0),
  RamanSettings_DeleteSpectraSection_StreamFind(list("shift" = c(-100, 350))),
  RamanSettings_DeleteSpectraSection_StreamFind(list("shift" = c(1800, 2500))),
  RamanSettings_CorrectSpectraBaseline_airpls(lambda = 45, differences = 1, itermax = 30)
)

raman_quant$add_settings(raman_workflow_settings)

raman_quant$print_workflow()
```

```{r raman-conc-run-workflow}
raman_quant$run_workflow()
```

```{r plot-raman-conc-spectra, fig.cap="Pre-processed Raman spectra for analysis replicate."}
raman_quant$plot_spectra()
```

## Applying MCR-ALS

The quantification of paracetamol from the Raman spectra is performed using the Multivariate Curve Resolution (MCR) with Alternating Least Squares (ALS) method via integration of the [mdatools](https://mdatools.com/docs/mcr--als.html) R package.

```{r mcrals-show}
stat3 <- StatisticEngine$new(data = raman_quant$get_spectra_matrix())
stat3$add_concentrations(c(1, 2.5, 5.0, 7.5, 15.0, NA_real_)) # add concentrations, with NA for unknown
stat3$process(StatisticSettings_MakeModel_mcrals_mdatools(ncomp = 3))
summary(stat3$model)
```

```{r mcrals-resolved-spec, fig.cap="Resolved spectra for each component and the spectra for each analysis replicate (i.e., each concentration)."}
stat3$plot_model_resolved_spectra()
```

```{r mcrals-cont, fig.cap="Contributions for each analysis replicate."}
stat3$plot_model_contributions()
```

The `quantify` method can be used to estimate the concentration of the paracetamol analyses with unknown concentration. The quantification is applied to each component, returning a list with the r square of the linear model and the estimated concentration as `data.table`. Note that the explained variance for each component is important to evaluate the calculated concentration. The best estimation is from the component with the combined explained variance and r square closer to 100 and 1, respectively.

```{r}
stat3$quantify()
```
